---
title: "shap_analysis (XAI for IMC Classifier)"
---

```{r load-shap-libraries, message=FALSE}
library(xgboost)
library(shapviz)
library(SHAPforxgboost)
library(ggplot2)
library(dplyr)
library(patchwork)
library(purrr)
library(SingleCellExperiment)
library(SummarizedExperiment)
```

Ensuring that base_path and objects from the previous step are accessible.
```{r}
base_path <- file.path(path.expand("~"), "Desktop", "proteomics", "XGBoost_classifier", "Panel_2_10")
final_model <- readRDS(file.path(base_path, "classifier_xgboost.rds"))
train_sce <- readRDS(file.path(base_path, "ROut", "train_sce.rds"))
```

Reconstruct the exact feature matrix X used as input for the final model
```{r}
marker_keep <- !grepl("DNA|Histone", rownames(train_sce))
X <- t(assay(train_sce, "exprs")[marker_keep, ])
```

Retrieve cell type names (class mapping)
```{r}
label_mapping <- levels(factor(train_sce$cell_labels))
```

Sanity check for feature matching
```{r}
stopifnot(ncol(X) == 41)
```

Compute SHAP Values via shapviz
```{r}
shp_bundle <- shapviz(final_model, X_pred = X)
names(shp_bundle) <- label_mapping
print(shp_bundle)
```

# Global SHAP & Structural Interpretations

## 1. Global mean |SHAP| bar chart
This averages absolute SHAP values across ALL cell classes to show overall marker ranking.
```{r global-and-structural-shap}
# Remove the "undefied" class from cell types
shp_bundle_clean <- mshapviz(shp_bundle[names(shp_bundle) != "undefined"])

p_global_imp <- sv_importance(shp_bundle_clean, kind = "bar", max_display = 20)
p_global_imp$mapping$fill <- aes(fill = Feature)$fill

distinct_20_palette <- c(
  "#2A4B7C", "#E2583E", "#398D8F", "#F2B705", 
  "#8E44AD", "#2ECC71", "#D35400", "#34495E", 
  "#1ABC9C", "#C0392B", "#27AE60", "#F39C12", 
  "#7F8C8D", "#2980B9", "#884EA0", "#D5F5E3", 
  "#E6B0AA", "#A9DFBF", "#F9E79F", "#D5D8DC"
)

p_global_imp_final <- p_global_imp +
  coord_flip() +  
  scale_fill_manual(values = distinct_20_palette, guide = "none") + 
  theme_bw() +
  labs(
    title = "Global Mean |SHAP| Feature Importance",
    subtitle = "Overall marker contribution averaged across all cell types",
    x = "Mean |SHAP Value|",
    y = "Markers"
  ) +
  theme(
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14)
  )

p_global_imp_final
ggsave(
  filename = file.path(base_path, "ROut", "shap_global_importance.png"), 
  plot = p_global_imp_final, 
  width = 16, 
  height = 9,
  dpi = 300
)
```

## 2. SHAP beeswarm — all test cells
This combines the distributions of all classes into a singular overarching summary plot.
```{r}
p_global_beeswarm <- sv_importance(shp_bundle_clean, kind = "beeswarm", max_display = 10) &
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 7, face = "bold"),
    axis.text.x = element_text(size = 8),
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 10)
  )

p_global_beeswarm
ggsave(
  filename = file.path(base_path, "ROut", "shap_beeswarm_global.png"), 
  plot = p_global_beeswarm, 
  width = 20, 
  height = 14,
  dpi = 300
)
```

## 3. Marker x cell-type SHAP heatmap
```{r}
mean_shap_df <- purrr::imap_dfr(shp_bundle_clean, function(shp, class_name) {
  data.frame(
    Marker = colnames(shp$S),
    Mean_Absolute_SHAP = colMeans(abs(shp$S)),
    Cell_Type = class_name
  )
})

p_heatmap <- ggplot(mean_shap_df, aes(x = Cell_Type, y = Marker, fill = Mean_Absolute_SHAP)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "inferno", name = "Mean |SHAP|") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Marker × Cell-Type SHAP Importance Heatmap",
    subtitle = "Direct visualization of learned feature importance per cell lineage",
    x = "Cell Phenotype Class",
    y = "Spatial Proteomics Marker"
  )

p_heatmap
ggsave(
  filename = file.path(base_path, "ROut", "shap_class_heatmap.png"), 
  plot = p_heatmap, 
  width = 16, 
  height = 9,
  dpi = 300
)

```

# Individual markers examinations (examples)

Function to generate and format the SHAP beeswarm plot
```{r}
plot_cell_type_shap <- function(shp_obj, cell_type_name, top_n = 15) {
  sv_importance(shp_obj[[cell_type_name]], kind = "beeswarm", max_display = top_n) +
    theme_bw() +
    labs(
      title = paste("SHAP Feature Contributions for Class:", cell_type_name),
      subtitle = "How marker expression drives the model choice for this cell type",
      x = "SHAP Value (Impact on Model Prediction)"
    )
}
```


## Feature contributions for some types of cells
```{r}

# Example A: Why does the model classify cells as CD8?
p_cd8 <- plot_cell_type_shap(shp_bundle, "CD8")
p_cd8

# Example B: What characterizes Tumor cells?
p_tumor <- plot_cell_type_shap(shp_bundle, "Tumor")
p_tumor

# Example C: How vCAF cells are classified?
p_vCAF <- plot_cell_type_shap(shp_bundle, "vCAF")
p_vCAF

#Save plots to the project output directory
ggsave(file.path(base_path, "ROut", "shap_beeswarm_CD8.png"), plot = p_cd8, width = 8, height = 6)
ggsave(file.path(base_path, "ROut", "shap_beeswarm_Tumor.png"), plot = p_tumor, width = 8, height = 6)
ggsave(file.path(base_path, "ROut", "shap_beeswarm_p_vCAF.png"), plot = p_tumor, width = 8, height = 6)
```

## Global feature importance for CD8 and NK
```{r}
#Global feature importance (Mean |SHAP|) for CD8
p_imp_cd8 <- sv_importance(shp_bundle$CD8, kind = "bar", max_display = 15) + 
  labs(title = "Feature Importance for CD8") + theme_minimal()

#Global feature importance (Mean |SHAP|) for NK
p_imp_nk <- sv_importance(shp_bundle$NK, kind = "bar", max_display = 15) + 
  labs(title = "Feature Importance for NK") + theme_minimal()

p_compare <- p_imp_cd8 + p_imp_nk
p_compare
ggsave(file.path(base_path, "ROut", "shap_importance_CD8_vs_NK.png"), plot = p_compare, width = 12, height = 6)
```

## Myeloid Differentiation: MacCD204 vs. MacCD209
We look at the top macrophage/dendritic subsets to understand how the model draws sharp boundaries between closely related myeloid phenotypes.
```{r shap-myeloid-dynamics, fig.height=6, fig.width=18}
p_mac204 <- plot_cell_type_shap(shp_bundle, "MacCD204", top_n = 10) +
  labs(subtitle = "Marker rules for M2-like Macrophages")

p_mac209 <- plot_cell_type_shap(shp_bundle, "MacCD209", top_n = 10) +
  labs(subtitle = "Marker rules for CD209+ Macrophages/DCs")

p_myeloid_compare <- p_mac204 + p_mac209
p_myeloid_compare

ggsave(file.path(base_path, "ROut", "shap_myeloid_comparison.png"), plot = p_myeloid_compare, width = 16, height = 6)
```

## CD3 x CD8a dependence plot
Analyzes the core lymphoid lineage switch to see how CD3 impact scales with CD8a expression.
```{r}
p_cd3_cd8a <- sv_dependence(
  shp_bundle$CD8, 
  v = "CD3", 
  color_var = "CD8a"
) +
  theme_bw() +
  labs(
    title = "Lymphoid Lineage Interaction: CD3 × CD8a",
    subtitle = "Deconstructing CD8 prediction logic across the CD3 and CD8a expression axes",
    x = "CD3 Expression Intensity (asinh)",
    y = "SHAP Value for CD8"
  )

p_cd3_cd8a
ggsave(
  filename = file.path(base_path, "ROut", "shap_interaction_CD3_CD8a.png"), 
  plot = p_cd3_cd8a, 
  width = 16, 
  height = 9
)
```

## Lymphoid Threshold: FOXP3 Activation in Tregs
```{r}
# Plot FOXP3 expression vs its SHAP value, color-coded by the lymphoid marker CD3
p_treg_threshold <- sv_dependence(
  shp_bundle$Treg, 
  v = "FOXP3", 
  color_var = "CD3"
) +
  theme_bw() +
  labs(
    title = "FOXP3 Activation Threshold for Treg Classification",
    subtitle = "Points above 0 show where FOXP3 confirms a Treg identity",
    x = "FOXP3 Expression Intensity (asinh)",
    y = "SHAP Value (Impact on Treg Prediction)"
  )

p_treg_threshold

ggsave(file.path(base_path, "ROut", "shap_treg_foxp3_threshold.png"), plot = p_treg_threshold, width = 16, height = 9)
```

## CD3 x CD20 dependence plot
```{r}
# Plot CD20 expression vs its SHAP value, color-coded by the lymphoid marker CD3
p_treg_threshold <- sv_dependence(
  shp_bundle$BnT, 
  v = "CD20", 
  color_var = "CD3"
) +
  theme_bw() +
  labs(
    title = "CD20 Activation Threshold for BnT Classification",
    x = "CD20 Expression Intensity (asinh)",
    y = "SHAP Value (Impact on BnT Prediction)"
  )

p_treg_threshold

ggsave(file.path(base_path, "ROut", "shap_BnT_CD20_threshold.png"), plot = p_treg_threshold, width = 16, height = 9)
```


## How CD146 expression shifts decision boundaries between vCAFs and SMA
```{r}
# SMA expression colored by CD146 (from vCAF perspective)
p_vcaf_cd146 <- shapviz::sv_dependence(
  shp_bundle$vCAF, 
  v = "SMA", 
  color_var = "CD146"
) + 
  theme_bw() + 
  labs(title = "vCAF Model Logic", x = "SMA Expression Intensity (asinh)", y = "SHAP for vCAF")


# SMA expression colored by CD146 (from SMA perspective)
p_sma_cd146 <- shapviz::sv_dependence(
  shp_bundle$SMA, 
  v = "SMA", 
  color_var = "CD146"
) + 
  theme_bw() + 
  labs(title = "SMA Model Logic", x = "SMA Expression Intensity (asinh)", y = "SHAP for SMA")

# Stitch CD146 interaction side-by-side
p_cd146_comparison <- (p_vcaf_cd146 | p_sma_cd146) + 
  plot_annotation(
    title = "Stromal Lineage Blur: SMA × CD146 Interaction",
    subtitle = "Analyzing how CD146 expression shifts decision boundaries between vCAFs and SMA"
  )

p_cd146_comparison

ggsave(
  filename = file.path(base_path, "ROut", "shap_blur_SMA_CD146.png"), 
  plot = p_cd146_comparison, 
  width = 16, 
  height = 7, 
  dpi = 300
)
```


