---
title: "03_uncertainty_analysis"
subtitle: "Model Uncertainty: Which Cells Should Not Be Trusted?"
author: "Inna Kucherova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
```

# Introduction

This notebook analyses the **uncertainty** of the XGBoost classifier.

The model does not simply assign a label — it outputs **softmax probabilities** for
each class. Cells with a uniform probability distribution are considered "ambiguous".

**Key metrics:**
- `max_prob` — confidence in the top predicted class
- `entropy` — prediction entropy H = -Σ p·log(p)

> "Which cells should not be trusted?"

---

# 0. Packages

```{r load-libraries}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  here, 
  conflicted,
  ggplot2, 
  dplyr, 
  tidyr, 
  forcats,
  patchwork, 
  viridis, 
  circlize, 
  ggridges, 
  RColorBrewer
)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("ComplexHeatmap", "SummarizedExperiment"), update = FALSE, ask = FALSE)

conflicted::conflicts_prefer(dplyr::count)
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(base::setdiff)
```

---

# 1. Load Objects

```{r load-objects}
base_path <- here::here("data", "Panel_2_10")

out_dir <- file.path(base_path, "ROut")

spe_27 <- readRDS(file.path(out_dir, "spe.rds"))

inputs        <- readRDS(file.path(out_dir, "classifier_inputs.rds"))
test_prob     <- inputs$test_prob
test_pred     <- inputs$test_pred
test_true     <- inputs$test_true
label_mapping <- inputs$label_mapping
test_sce      <- inputs$test_sce

num_classes <- length(label_mapping)
n_test      <- nrow(test_prob)

cat("Test cells:", n_test, "\n")
cat("Classes:", num_classes, "\n")
```

---

# 2. Compute Uncertainty Metrics

```{r compute-uncertainty}
# Maximum probability: the higher, the more confident the model
max_prob <- apply(test_prob, 1, max)

# Entropy: H = -Σ p·log2(p)  (0 = full confidence, log2(K) = maximum)
entropy <- apply(test_prob, 1, function(p) {
  p_pos <- p[p > 0]
  -sum(p_pos * log2(p_pos))
})

# Maximum possible entropy (uniform distribution)
max_entropy <- log2(num_classes)

# Normalized entropy [0, 1]
entropy_norm <- entropy / max_entropy

# Labels: predicted class and correctness
pred_class  <- factor(as.character(test_pred), levels = label_mapping)
true_class  <- factor(as.character(test_true), levels = label_mapping)
is_correct  <- pred_class == true_class

# Second best class and probability margin
second_prob <- apply(test_prob, 1, function(p) sort(p, decreasing = TRUE)[2])
prob_margin <- max_prob - second_prob

uncertainty_df <- data.frame(
  cell_id      = rownames(test_prob),
  max_prob     = max_prob,
  second_prob  = second_prob,
  prob_margin  = prob_margin,
  entropy      = entropy,
  entropy_norm = entropy_norm,
  pred_class   = pred_class,
  true_class   = true_class,
  is_correct   = is_correct,
  row.names    = NULL
)

cat("Threshold statistics (max_prob):\n")
cat("  < 0.5 (low confidence):", sum(max_prob < 0.5), "\n")
cat("  < 0.6:", sum(max_prob < 0.6), "\n")
cat("  < 0.8:", sum(max_prob < 0.8), "\n")
cat("  ≥ 0.9 (high confidence):", sum(max_prob >= 0.9), "\n")
```

---

# 3. Distribution of max_prob and Entropy

```{r distribution-plots, fig.width=12, fig.height=5}
p1 <- ggplot(uncertainty_df, aes(x = max_prob, fill = is_correct)) +
  geom_histogram(bins = 60, alpha = 0.8, position = "stack") +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = "#4575B4", "FALSE" = "#D73027"),
    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"),
    name = NULL
  ) +
  theme_classic(base_size = 13) +
  labs(
    title    = "Distribution of Maximum Probability",
    subtitle = "Red line = uncertainty threshold (max_prob < 0.6)",
    x        = "max softmax P",
    y        = "Cell count"
  )

p2 <- ggplot(uncertainty_df, aes(x = entropy_norm, fill = is_correct)) +
  geom_histogram(bins = 60, alpha = 0.8, position = "stack") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = "#4575B4", "FALSE" = "#D73027"),
    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"),
    name = NULL
  ) +
  theme_classic(base_size = 13) +
  labs(
    title    = "Distribution of Normalised Prediction Entropy",
    subtitle = "H = -Σ p·log₂(p), normalised to [0, 1]",
    x        = "Normalised entropy",
    y        = "Cell count"
  )

p1 + p2
```

```{r save-distribution}
ggsave(
  file.path(out_dir, "uncertainty_distributions.png"),
  p1 + p2, width = 14, height = 6, dpi = 150
)
```

---

# 4. Uncertainty per Cell Type

```{r uncertainty-per-class, fig.height=8, fig.width=10}
# Which cell types are present in the data?
cell_types <- levels(fct_reorder(uncertainty_df$pred_class, 
                                  uncertainty_df$entropy_norm, median))
cell_types
length(cell_types)

colors <- c(
  # Tumor
  "Tumor"       = "#E41A1C",
  
  # T cells
  "CD8"         = "#377EB8",
  "CD4"         = "#4DAF4A",
  "Treg"        = "#984EA3",
  "BnT"         = "#A65628",
  
  # NK, B cells
  "NK"          = "#FF7F00",
  "B"           = "#FFFF33",
  
  # Macrophages / dendritic cells
  "Mac"         = "#F781BF",
  "MacCD204"    = "#E7298A",
  "MacCD209"    = "#CE1256",
  "DC"          = "#66C2A5",
  "mregDC"      = "#3288BD",
  
  # MDSC / neutrophils
  "MDSC"        = "#FC8D59",
  "PMN_MDSC"    = "#D53E4F",
  "Neutrophil"  = "#FEE08B",
  
  # Stroma / vasculature
  "Fibro"       = "#8DA0CB",
  "vCAF"        = "#A6D854",
  "SMA"         = "#FFD92F",
  "Vasculature" = "#E5C494",
  
  # Other
  "Igg"         = "#B3B3B3",
  "undefined"   = "#CCCCCC"
)

# Check that all cell types are covered
base::setdiff(cell_types, names(colors)) # should return character(0)

# Save to metadata
metadata(spe_27)$colour_vectors$cell_types <- colors

# Build the plot
p_ridge <- ggplot(
  uncertainty_df,
  aes(x = entropy_norm, y = fct_reorder(pred_class, entropy_norm, median), fill = pred_class)
) +
  geom_density_ridges(alpha = 0.75, scale = 0.9, rel_min_height = 0.01) +
  scale_fill_manual(values = colors, guide = "none") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red") +
  theme_classic(base_size = 12) +
  labs(
    title    = "Prediction Entropy by Cell Type",
    subtitle = "Ordered by median entropy",
    x        = "Normalised entropy",
    y        = "Predicted cell type"
  )
p_ridge
```

```{r save-ridge}
ggsave(
  file.path(out_dir, "uncertainty_ridgeplot.png"),
  p_ridge, width = 10, height = 8, dpi = 150
)
```

---

# 5. Low-Confidence Cells (Uncertain Cells)

```{r uncertain-cells}
CONFIDENCE_THRESHOLD <- 0.6

uncertain_df <- uncertainty_df |> 
  dplyr::filter(max_prob < CONFIDENCE_THRESHOLD)

cat("Uncertain cells (max_prob <", CONFIDENCE_THRESHOLD, "):",
    nrow(uncertain_df), "/", n_test, 
    "(", round(100 * nrow(uncertain_df) / n_test, 1), "%)\n")

# Breakdown by predicted class
uncertain_by_class <- uncertain_df |>
  dplyr::count(pred_class, name = "n_uncertain") |>
  dplyr::left_join(
    uncertainty_df |> dplyr::count(pred_class, name = "n_total"),
    by = "pred_class"
  ) |>
  dplyr::mutate(pct_uncertain = round(100 * n_uncertain / n_total, 1)) |>
  dplyr::arrange(dplyr::desc(pct_uncertain))

uncertain_by_class
```

```{r plot-uncertain-per-class, fig.height=6, fig.width=9}
p_unc <- ggplot(uncertain_by_class, aes(
  x    = pct_uncertain,
  y    = fct_reorder(pred_class, pct_uncertain),
  fill = pred_class
)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(
    values = metadata(spe_27)$colour_vectors$cell_types,
    guide  = "none"
  ) +
  geom_text(
    aes(label = paste0(pct_uncertain, "%")),
    hjust = -0.1, size = 3.5
  ) +
  xlim(0, max(uncertain_by_class$pct_uncertain) * 1.2) +
  theme_classic(base_size = 12) +
  labs(
    title    = "% Uncertain Cells per Cell Type (max_prob < 0.6)",
    subtitle = "Classes with a higher uncertain fraction are harder for the classifier",
    x        = "% uncertain",
    y        = "Predicted cell type"
  )

p_unc
```

```{r save-uncertain-plot}
ggsave(
  file.path(out_dir, "uncertainty_per_class.png"),
  p_unc, width = 9, height = 6, dpi = 150
)
```

---

# 6. Confusion Matrix for Uncertain Cells Only

```{r confusion-uncertain, fig.height=8, fig.width=10}
library(caret)

# Exclude NA in true_class
uncertain_labeled <- uncertain_df |> 
  dplyr::filter(!is.na(true_class))

if (nrow(uncertain_labeled) > 10) {
  cm_unc <- caret::confusionMatrix(
    data      = factor(uncertain_labeled$pred_class, levels = label_mapping),
    reference = factor(uncertain_labeled$true_class,  levels = label_mapping)
  )
  
  cm_df <- as.data.frame(cm_unc$table)
  
  p_cm_unc <- ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "#D73027", name = "Cell count") +
    geom_text(aes(label = ifelse(Freq > 0, Freq, "")), size = 2.5) +
    theme_classic(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(size = 9)
    ) +
    labs(
      title    = "Confusion Matrix: Uncertain Cells Only (max_prob < 0.6)",
      subtitle = paste("Accuracy:", round(cm_unc$overall["Accuracy"] * 100, 1), "%"),
      x        = "True class",
      y        = "Predicted class"
    )
  
  p_cm_unc
  
  ggsave(
    file.path(out_dir, "uncertainty_confusion_matrix.png"),
    p_cm_unc, width = 12, height = 10, dpi = 150
  )
}
```

---

# 7. Most Confused Class Pairs

```{r confusion-pairs, fig.height=6, fig.width=10}
# Extract incorrect predictions from all test cells
error_df <- uncertainty_df |> 
  dplyr::filter(!is_correct, !is.na(true_class)) |> 
  dplyr::mutate(
    confusion_pair = paste0(
      as.character(true_class), " → ", as.character(pred_class)
    )
  ) |> 
  dplyr::count(
    true_class, pred_class, confusion_pair,
    name = "n_errors"
  ) |> 
  dplyr::arrange(dplyr::desc(n_errors)) |> 
  head(20)

p_errors <- ggplot(error_df, aes(
  x    = n_errors,
  y    = fct_reorder(confusion_pair, n_errors),
  fill = n_errors
)) +
  geom_col(alpha = 0.9) +
  scale_fill_gradient(low = "#FEE090", high = "#D73027", guide = "none") +
  geom_text(aes(label = n_errors), hjust = -0.2, size = 3.5) +
  xlim(0, max(error_df$n_errors) * 1.2) +
  theme_classic(base_size = 12) +
  labs(
    title    = "Top-20 Confusion Pairs (All Test Cells)",
    subtitle = "Format: True class → Predicted class",
    x        = "Number of errors",
    y        = NULL
  )

p_errors
```

```{r save-confusion-pairs}
ggsave(
  file.path(out_dir, "uncertainty_confusion_pairs.png"),
  p_errors, width = 10, height = 7, dpi = 150
)
```

---

# 8. Entropy vs. Margin: Scatter Plot

```{r entropy-margin-scatter, fig.height=6, fig.width=9}
p_scatter <- ggplot(
  uncertainty_df |> dplyr::filter(!is.na(true_class)),
  aes(x = entropy_norm, y = prob_margin, color = is_correct)
) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_manual(
    values = c("TRUE" = "#4575B4", "FALSE" = "#D73027"),
    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"),
    name   = NULL
  ) +
  geom_hline(yintercept = 0.3, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "gray50") +
  theme_classic(base_size = 13) +
  labs(
    title    = "Entropy vs. Probability Margin",
    subtitle = "Misclassified cells: high entropy + small margin",
    x        = "Normalised entropy H",
    y        = "P(1st) − P(2nd) = margin"
  )

p_scatter
```

```{r save-scatter}
ggsave(
  file.path(out_dir, "uncertainty_entropy_margin_scatter.png"),
  p_scatter, width = 9, height = 6, dpi = 150
)
```

---

# 9. Uncertainty Summary Table

```{r summary-table}
summary_unc <- uncertainty_df |> 
  dplyr::filter(!is.na(true_class)) |> 
  dplyr::group_by(pred_class) |> 
  dplyr::summarise(
    n_total        = n(),
    n_uncertain    = sum(max_prob < CONFIDENCE_THRESHOLD),
    pct_uncertain  = round(100 * n_uncertain / n_total, 1),
    median_entropy = round(median(entropy_norm), 3),
    median_maxprob = round(median(max_prob), 3),
    accuracy       = round(mean(is_correct) * 100, 1),
    .groups = "drop"
  ) |> 
  dplyr::arrange(dplyr::desc(pct_uncertain))

summary_unc

write.csv(
  summary_unc,
  file.path(out_dir, "uncertainty_summary_per_class.csv"),
  row.names = FALSE
)
```

---

# Conclusions

1. Cells with **max_prob < 0.6** (≈ X% of the test set) are ambiguous —
   the model cannot confidently assign a single type.
2. **Highest entropy** is observed in rare or transitional classes:
   Igg, PMN_MDSC, Neutrophil, MDSC — consistent with low accuracy in the confusion matrix.
3. **Most frequent confusion pairs** reflect biological similarity:
   MacCD204 ↔ Mac, CD4 ↔ CD8, Fibro ↔ SMA.
4. **High entropy + small margin** → a cell is a true "borderline" case;
   it should be treated with particular caution in downstream analyses.
