---
title: "04_uncertainty_comparison"
subtitle: "Model Uncertainty: Test Set (6k) vs. Full Dataset (250k)"
author: "Inna Kucherova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 7)
```

# Introduction

This notebook compares XGBoost classifier uncertainty across two populations:

- **Test set (6k cells)** — 27 manually gated images, true labels available → correctness known  
- **Unseen set (250k cells)** — ~60 images never seen during training, no true labels

**Key metrics:**  
- `max_prob` — confidence in the top predicted class  
- `entropy` — prediction entropy H = −Σ p·log₂(p), normalised to [0, 1]

> "Does the model behave consistently on new data?"

---

# 0. Packages

```{r load-libraries}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  here, conflicted,
  ggplot2, dplyr, tidyr, forcats,
  patchwork, viridis, ggridges,
  SingleCellExperiment, SummarizedExperiment
)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("ComplexHeatmap", "SummarizedExperiment"),
                     update = FALSE, ask = FALSE)

conflicted::conflicts_prefer(dplyr::count)
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(base::setdiff)
```

---

# 1. Load Objects

```{r load-objects}
base_path <- here::here("data", "Panel_2_10")
out_dir <- file.path(base_path, "ROut")

# ── Test set (6k) ──────────────────────────────────────────────────────────
inputs        <- readRDS(file.path(out_dir, "classifier_inputs.rds"))
test_prob     <- inputs$test_prob       # matrix: cells × classes
test_pred     <- inputs$test_pred       # factor: predicted labels
test_true     <- inputs$test_true       # factor: true labels
label_mapping <- inputs$label_mapping   # character vector of class names
test_sce      <- inputs$test_sce

num_classes <- length(label_mapping)

cat("=== Test set ===\n")
cat("Cells:", nrow(test_prob), "| Classes:", num_classes, "\n\n")

# ── Unseen set (250k) ──────────────────────────────────────────────────────
# IMPORTANT: spe_unseen_predicted.rds must contain $prediction_entropy_norm
# and $prediction_confidence. Re-run predict_unlabeled.Rmd with the
# entropy patch before executing this notebook.
spe_unseen <- readRDS(file.path(out_dir, "spe_unseen_predicted.rds"))

# Load the full probability matrix saved by predict_unlabeled.Rmd
unseen_prob <- readRDS(file.path(out_dir, "unseen_prob_matrix.rds"))

cat("=== Unseen set ===\n")
cat("Cells:", ncol(spe_unseen), "\n")
```

---

# 2. Build Unified Uncertainty Data Frames

```{r compute-uncertainty-test}
# ── Test set ────────────────────────────────────────────────────────────────
max_prob_test  <- apply(test_prob, 1, max)
second_test    <- apply(test_prob, 1, function(p) sort(p, decreasing = TRUE)[2])
entropy_test   <- apply(test_prob, 1, function(p) { p <- p[p > 0]; -sum(p * log2(p)) })
entropy_norm_t <- entropy_test / log2(num_classes)

pred_class_t   <- factor(as.character(test_pred), levels = label_mapping)
true_class_t   <- factor(as.character(test_true), levels = label_mapping)

unc_test <- data.frame(
  cell_id      = rownames(test_prob),
  max_prob     = max_prob_test,
  second_prob  = second_test,
  prob_margin  = max_prob_test - second_test,
  entropy      = entropy_test,
  entropy_norm = entropy_norm_t,
  pred_class   = pred_class_t,
  true_class   = true_class_t,
  is_correct   = pred_class_t == true_class_t,
  dataset      = "Test (6k)",
  row.names    = NULL
)
```

```{r compute-uncertainty-unseen}
# ── Unseen set ──────────────────────────────────────────────────────────────
# Use the saved prob matrix to recompute metrics consistently
max_prob_u   <- apply(unseen_prob, 1, max)
second_u     <- apply(unseen_prob, 1, function(p) sort(p, decreasing = TRUE)[2])
entropy_u    <- apply(unseen_prob, 1, function(p) { p <- p[p > 0]; -sum(p * log2(p)) })
entropy_norm_u <- entropy_u / log2(num_classes)

pred_class_u <- factor(
  as.character(spe_unseen$predicted_label),
  levels = label_mapping
)

unc_unseen <- data.frame(
  cell_id      = colnames(spe_unseen),
  max_prob     = max_prob_u,
  second_prob  = second_u,
  prob_margin  = max_prob_u - second_u,
  entropy      = entropy_u,
  entropy_norm = entropy_norm_u,
  pred_class   = pred_class_u,
  true_class   = NA_character_,   # no ground truth
  is_correct   = NA,
  dataset      = "Unseen (250k)",
  row.names    = NULL
)
```

```{r combine}
# Combined data frame for side-by-side comparisons
unc_all <- dplyr::bind_rows(unc_test, unc_unseen) |>
  dplyr::mutate(dataset = factor(dataset, levels = c("Test (6k)", "Unseen (250k)")))

cat("Combined rows:", nrow(unc_all), "\n")
```

---

# 3. Colour Palette

```{r colours}
colors <- c(
  "Tumor"       = "#E41A1C",
  "CD8"         = "#377EB8",
  "CD4"         = "#4DAF4A",
  "Treg"        = "#984EA3",
  "BnT"         = "#A65628",
  "NK"          = "#FF7F00",
  "B"           = "#FFFF33",
  "Mac"         = "#F781BF",
  "MacCD204"    = "#E7298A",
  "MacCD209"    = "#CE1256",
  "DC"          = "#66C2A5",
  "mregDC"      = "#3288BD",
  "MDSC"        = "#FC8D59",
  "PMN_MDSC"    = "#D53E4F",
  "Neutrophil"  = "#FEE08B",
  "Fibro"       = "#8DA0CB",
  "vCAF"        = "#A6D854",
  "SMA"         = "#FFD92F",
  "Vasculature" = "#E5C494",
  "Igg"         = "#B3B3B3",
  "undefined"   = "#CCCCCC"
)
```

---

# 4. Distribution of max_prob and Entropy

The test set can be split by correctness; the unseen set cannot.

```{r distribution-test, fig.width=12, fig.height=5}
p_prob_t <- ggplot(unc_test, aes(x = max_prob, fill = is_correct)) +
  geom_histogram(bins = 60, alpha = 0.85, position = "stack") +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = "#4575B4", "FALSE" = "#D73027"),
    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"), name = NULL
  ) +
  theme_classic(base_size = 13) +
  labs(title = "Test set (6k) — max_prob",
       subtitle = "Red line = threshold 0.6",
       x = "max softmax P", y = "Cell count")

p_entr_t <- ggplot(unc_test, aes(x = entropy_norm, fill = is_correct)) +
  geom_histogram(bins = 60, alpha = 0.85, position = "stack") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = "#4575B4", "FALSE" = "#D73027"),
    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"), name = NULL
  ) +
  theme_classic(base_size = 13) +
  labs(title = "Test set (6k) — normalised entropy",
       subtitle = "H = −Σ p·log₂(p), normalised to [0, 1]",
       x = "Normalised entropy", y = "Cell count")

p_prob_t + p_entr_t
```

```{r distribution-unseen, fig.width=12, fig.height=5}
p_prob_u <- ggplot(unc_unseen, aes(x = max_prob)) +
  geom_histogram(bins = 60, alpha = 0.85, fill = "#4575B4") +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "red", linewidth = 0.8) +
  theme_classic(base_size = 13) +
  labs(title = "Unseen set (250k) — max_prob",
       subtitle = "No ground truth available",
       x = "max softmax P", y = "Cell count")

p_entr_u <- ggplot(unc_unseen, aes(x = entropy_norm)) +
  geom_histogram(bins = 60, alpha = 0.85, fill = "#4575B4") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red", linewidth = 0.8) +
  theme_classic(base_size = 13) +
  labs(title = "Unseen set (250k) — normalised entropy",
       x = "Normalised entropy", y = "Cell count")

p_prob_u + p_entr_u
```

```{r distribution-overlay, fig.width=12, fig.height=5}
# Density overlay: how similar are the two distributions?
p_prob_ov <- ggplot(unc_all, aes(x = max_prob, colour = dataset, fill = dataset)) +
  geom_density(alpha = 0.25, linewidth = 0.9) +
  geom_vline(xintercept = 0.6, linetype = "dashed", color = "red") +
  scale_colour_manual(values = c("Test (6k)" = "#D73027", "Unseen (250k)" = "#4575B4")) +
  scale_fill_manual(values  = c("Test (6k)" = "#D73027", "Unseen (250k)" = "#4575B4")) +
  theme_classic(base_size = 13) +
  labs(title = "max_prob — Test vs. Unseen (density overlay)",
       x = "max softmax P", y = "Density", colour = NULL, fill = NULL)

p_entr_ov <- ggplot(unc_all, aes(x = entropy_norm, colour = dataset, fill = dataset)) +
  geom_density(alpha = 0.25, linewidth = 0.9) +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red") +
  scale_colour_manual(values = c("Test (6k)" = "#D73027", "Unseen (250k)" = "#4575B4")) +
  scale_fill_manual(values  = c("Test (6k)" = "#D73027", "Unseen (250k)" = "#4575B4")) +
  theme_classic(base_size = 13) +
  labs(title = "Entropy — Test vs. Unseen (density overlay)",
       x = "Normalised entropy", y = "Density", colour = NULL, fill = NULL)

p_prob_ov + p_entr_ov
```

```{r save-distributions}
ggsave(file.path(out_dir, "comp_distributions_test.png"),
       p_prob_t + p_entr_t, width = 14, height = 6, dpi = 150)
ggsave(file.path(out_dir, "comp_distributions_unseen.png"),
       p_prob_u + p_entr_u, width = 14, height = 6, dpi = 150)
ggsave(file.path(out_dir, "comp_distributions_overlay.png"),
       p_prob_ov + p_entr_ov, width = 14, height = 6, dpi = 150)
```

---

# 5. Entropy per Cell Type (Ridge Plots)

```{r ridge-test, fig.height=9, fig.width=10}
p_ridge_t <- ggplot(
  unc_test,
  aes(x = entropy_norm,
      y = fct_reorder(pred_class, entropy_norm, median),
      fill = pred_class)
) +
  geom_density_ridges(alpha = 0.75, scale = 0.9, rel_min_height = 0.01) +
  scale_fill_manual(values = colors, guide = "none") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red") +
  theme_classic(base_size = 12) +
  labs(title = "Test set (6k) — Entropy by Cell Type",
       subtitle = "Ordered by median entropy",
       x = "Normalised entropy", y = NULL)

p_ridge_t
```

```{r ridge-unseen, fig.height=9, fig.width=10}
p_ridge_u <- ggplot(
  unc_unseen,
  aes(x = entropy_norm,
      y = fct_reorder(pred_class, entropy_norm, median),
      fill = pred_class)
) +
  geom_density_ridges(alpha = 0.75, scale = 0.9, rel_min_height = 0.01) +
  scale_fill_manual(values = colors, guide = "none") +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "red") +
  theme_classic(base_size = 12) +
  labs(title = "Unseen set (250k) — Entropy by Cell Type",
       subtitle = "Ordered by median entropy",
       x = "Normalised entropy", y = NULL)

p_ridge_u
```

```{r save-ridges}
ggsave(file.path(out_dir, "comp_ridge_test.png"),
       p_ridge_t, width = 10, height = 9, dpi = 150)
ggsave(file.path(out_dir, "comp_ridge_unseen.png"),
       p_ridge_u, width = 10, height = 9, dpi = 150)
```

---

# 6. % Uncertain Cells per Cell Type

```{r uncertain-cells}
CONFIDENCE_THRESHOLD <- 0.6

compute_uncertain <- function(df) {
  df |>
    dplyr::count(pred_class, name = "n_total") |>
    dplyr::left_join(
      df |>
        dplyr::filter(max_prob < CONFIDENCE_THRESHOLD) |>
        dplyr::count(pred_class, name = "n_uncertain"),
      by = "pred_class"
    ) |>
    dplyr::mutate(
      n_uncertain  = tidyr::replace_na(n_uncertain, 0),
      pct_uncertain = round(100 * n_uncertain / n_total, 1)
    ) |>
    dplyr::arrange(dplyr::desc(pct_uncertain))
}

unc_by_class_t <- compute_uncertain(unc_test)
unc_by_class_u <- compute_uncertain(unc_unseen)

cat("=== Test set: uncertain cells (max_prob < 0.6) ===\n")
cat(sum(unc_test$max_prob < CONFIDENCE_THRESHOLD), "/", nrow(unc_test),
    "(", round(100 * mean(unc_test$max_prob < CONFIDENCE_THRESHOLD), 1), "%)\n\n")

cat("=== Unseen set: uncertain cells (max_prob < 0.6) ===\n")
cat(sum(unc_unseen$max_prob < CONFIDENCE_THRESHOLD), "/", nrow(unc_unseen),
    "(", round(100 * mean(unc_unseen$max_prob < CONFIDENCE_THRESHOLD), 1), "%)\n")
```

```{r plot-uncertain-comparison, fig.width=14, fig.height=7}
plot_uncertain_bar <- function(df, title) {
  ggplot(df, aes(
    x    = pct_uncertain,
    y    = fct_reorder(pred_class, pct_uncertain),
    fill = pred_class
  )) +
    geom_col(alpha = 0.85) +
    scale_fill_manual(values = colors, guide = "none") +
    geom_text(aes(label = paste0(pct_uncertain, "%")),
              hjust = -0.1, size = 3.5) +
    xlim(0, max(df$pct_uncertain, na.rm = TRUE) * 1.2) +
    theme_classic(base_size = 12) +
    labs(title = title,
         x = "% uncertain (max_prob < 0.6)", y = NULL)
}

p_unc_t <- plot_uncertain_bar(unc_by_class_t, "Test set (6k)")
p_unc_u <- plot_uncertain_bar(unc_by_class_u, "Unseen set (250k)")

p_unc_t + p_unc_u
```

```{r save-uncertain-bars}
ggsave(file.path(out_dir, "comp_uncertain_bars.png"),
       p_unc_t + p_unc_u, width = 16, height = 8, dpi = 150)
```

---

# 7. Entropy vs. Margin Scatter

For the test set, points are coloured by correctness. For the unseen set, a single colour is used.

```{r scatter-test, fig.width=9, fig.height=6}
p_sc_t <- ggplot(
  unc_test,
  aes(x = entropy_norm, y = prob_margin, colour = is_correct)
) +
  geom_point(alpha = 0.35, size = 0.7) +
  scale_colour_manual(
    values = c("TRUE" = "#4575B4", "FALSE" = "#D73027"),
    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"), name = NULL
  ) +
  geom_hline(yintercept = 0.3, linetype = "dashed", colour = "gray50") +
  geom_vline(xintercept = 0.3, linetype = "dashed", colour = "gray50") +
  theme_classic(base_size = 13) +
  labs(title = "Test set (6k) — Entropy vs. Margin",
       subtitle = "Misclassified: high entropy + small margin",
       x = "Normalised entropy H", y = "P(1st) − P(2nd)")

p_sc_t
```

```{r scatter-unseen, fig.width=9, fig.height=6}
# Downsample for faster rendering (optional — remove if not needed)
set.seed(42)
unc_unseen_sample <- unc_unseen |> dplyr::slice_sample(n = min(50000, nrow(unc_unseen)))

p_sc_u <- ggplot(
  unc_unseen_sample,
  aes(x = entropy_norm, y = prob_margin)
) +
  geom_point(alpha = 0.15, size = 0.5, colour = "#4575B4") +
  geom_hline(yintercept = 0.3, linetype = "dashed", colour = "gray50") +
  geom_vline(xintercept = 0.3, linetype = "dashed", colour = "gray50") +
  theme_classic(base_size = 13) +
  labs(title = "Unseen set (250k) — Entropy vs. Margin",
       subtitle = "50k random cells shown",
       x = "Normalised entropy H", y = "P(1st) − P(2nd)")

p_sc_u
```

```{r save-scatter}
ggsave(file.path(out_dir, "comp_scatter_test.png"),
       p_sc_t, width = 9, height = 6, dpi = 150)
ggsave(file.path(out_dir, "comp_scatter_unseen.png"),
       p_sc_u, width = 9, height = 6, dpi = 150)
```

---

# 8. Confusion Pairs — Test Set Only

> This section is only possible where true labels exist (test set).

```{r confusion-pairs, fig.height=6, fig.width=10}
error_df <- unc_test |>
  dplyr::filter(!is_correct, !is.na(true_class)) |>
  dplyr::mutate(
    confusion_pair = paste0(as.character(true_class), " → ", as.character(pred_class))
  ) |>
  dplyr::count(true_class, pred_class, confusion_pair, name = "n_errors") |>
  dplyr::arrange(dplyr::desc(n_errors)) |>
  head(20)

p_errors <- ggplot(error_df, aes(
  x    = n_errors,
  y    = fct_reorder(confusion_pair, n_errors),
  fill = n_errors
)) +
  geom_col(alpha = 0.9) +
  scale_fill_gradient(low = "#FEE090", high = "#D73027", guide = "none") +
  geom_text(aes(label = n_errors), hjust = -0.2, size = 3.5) +
  xlim(0, max(error_df$n_errors) * 1.2) +
  theme_classic(base_size = 12) +
  labs(title = "Top-20 Confusion Pairs — Test Set",
       subtitle = "Format: True class → Predicted class",
       x = "Number of errors", y = NULL)

p_errors
```

```{r save-confusion-pairs}
ggsave(file.path(out_dir, "comp_confusion_pairs_test.png"),
       p_errors, width = 10, height = 7, dpi = 150)
```

---

# 9. Summary Tables

```{r summary-table-test}
summary_test <- unc_test |>
  dplyr::group_by(pred_class) |>
  dplyr::summarise(
    n_total        = n(),
    n_uncertain    = sum(max_prob < CONFIDENCE_THRESHOLD),
    pct_uncertain  = round(100 * n_uncertain / n_total, 1),
    median_entropy = round(median(entropy_norm), 3),
    median_maxprob = round(median(max_prob), 3),
    accuracy       = round(mean(is_correct, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(pct_uncertain))

cat("=== Test set summary ===\n")
print(summary_test)

write.csv(summary_test,
          file.path(out_dir, "comp_summary_test.csv"),
          row.names = FALSE)
```

```{r summary-table-unseen}
summary_unseen <- unc_unseen |>
  dplyr::group_by(pred_class) |>
  dplyr::summarise(
    n_total        = n(),
    n_uncertain    = sum(max_prob < CONFIDENCE_THRESHOLD),
    pct_uncertain  = round(100 * n_uncertain / n_total, 1),
    median_entropy = round(median(entropy_norm), 3),
    median_maxprob = round(median(max_prob), 3),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(pct_uncertain))

cat("=== Unseen set summary ===\n")
print(summary_unseen)

write.csv(summary_unseen,
          file.path(out_dir, "comp_summary_unseen.csv"),
          row.names = FALSE)
```

```{r summary-joint}
# Join both summaries for a side-by-side table
summary_joint <- dplyr::full_join(
  summary_test   |> dplyr::select(pred_class, pct_uncertain, median_entropy, median_maxprob, accuracy),
  summary_unseen |> dplyr::select(pred_class, pct_uncertain, median_entropy, median_maxprob),
  by = "pred_class", suffix = c("_test", "_unseen")
) |>
  dplyr::arrange(dplyr::desc(pct_uncertain_test))

summary_joint

write.csv(summary_joint,
          file.path(out_dir, "comp_summary_joint.csv"),
          row.names = FALSE)
```

---

# Conclusions

1. **Distribution shift confirmed**: the density overlay appears similar at first
   glance, but the unseen set histogram reveals a distinct second peak at entropy ≈ 1.0 and a
   low-confidence tail at max_prob 0–0.25, absent in the test set. The model is substantially
   less certain on the 60 new images than on the 27 it was trained on.

2. **Bimodal entropy in the unseen set**: unlike the test set — where all cell
   types show a unimodal distribution concentrated near H = 0 — many types in the unseen set
   display bimodal ridge distributions (NK, Igg, SMA, MDSC, DC, mregDC, MacCD209, B, CD4).
   This indicates two sub-populations within each type: cells the model recognises confidently,
   and cells it cannot resolve at all.

3. **Uncertain fraction — dramatic scale-up**: in the test set only NK (33.3%) and
   Neutrophil (21.4%) exceed the 20% flag threshold. In the unseen set, 14 out of 21 types
   exceed it, with NK reaching 90.7% and Igg 81.5%. NK and Igg predictions in downstream
   analyses of the unseen set should be treated as unreliable. Tumor (6.9%) and BnT (2.6%)
   remain the most stable types across both populations.

4. **Entropy vs. margin — geometry changes**: the test set scatter forms a compact
   triangle in the low-entropy / high-margin zone. The unseen set scatter stretches along the
   full diagonal to H = 1.0, margin = 0, with the bottom-right quadrant (high entropy + small
   margin) densely populated — confirming a large pool of genuinely unresolvable cells.

5. **Confusion pairs** (test set only — no ground truth for unseen): the dominant
   error is **vCAF → SMA (15 cells, 41% of all errors)**, driven by co-expression of α-SMA in
   both types. **Fibro → SMA (5 cells)** reflects the same stromal ambiguity. Minor pairs
   include MDSC → Neutrophil (2) and CD4 → CD8 (2), consistent with granulocytic and T-cell
   phenotypic overlap respectively. Note: Mac ↔ MacCD204 confusion was not observed in this
   test set.
