---
title: "04_spatial_xai"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This notebook asks whether **spatial context** influences cell identity — and whether the XGBoost classifier implicitly captures it.

Questions answered:

1. Which cell-type pairs are spatially enriched or depleted relative to chance?
2. Do cells with the same marker expression but different neighbourhoods receive different labels?
3. Is CD8+ T cell proximity to Tumour cells detectable at the population level?
4. How much does adding neighbourhood composition as features change classifier predictions?

Technically: `imcRtools::buildSpatialGraph()` → `countInteractions()` → `testInteractions()` + neighbourhood feature augmentation.

---

## Libraries

```{r libraries}
library(imcRtools)
library(SpatialExperiment)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(xgboost)
```

---

## Load labelled SPE

Loads the compensated, labelled SPE produced by `01_read_data.Rmd`.
Only cells with a confirmed label are kept — unlabelled cells have no ground truth for interaction analysis.

```{r load-spe}
base_path <- file.path(
  path.expand("~"),
  "Desktop",
  "KSE_Bioinformatics",
  "Protomics",
  "XGBoost_classifier",
  "XGBoost_classifier",
  "Panel_2_10"
)

plots_dir <- file.path(dirname(base_path), "plots")
dir.create(plots_dir, showWarnings = FALSE, recursive = TRUE)

spe <- readRDS(file.path(base_path, "ROut", "spe_labeled.rds"))

# drop residual NAs
spe <- spe[, !is.na(spe$cell_labels)]

# drop images with too few labelled cells for meaningful spatial analysis
# (images with < 50 labelled cells are manual-gate slivers — not spatially informative)
cell_counts   <- table(spe$sample_id)
keep_imgs     <- names(cell_counts[cell_counts >= 50])
dropped_imgs  <- names(cell_counts[cell_counts < 50])
if (length(dropped_imgs)) {
  cat("Dropping", length(dropped_imgs), "images with < 50 labelled cells:\n")
  print(sort(cell_counts[dropped_imgs]))
}
spe <- spe[, spe$sample_id %in% keep_imgs]
spe$sample_id <- droplevels(factor(spe$sample_id))

cat("\nCells:", ncol(spe), "\n")
cat("Images:", length(unique(spe$sample_id)), "\n")
cat("Cell types:\n")
print(table(spe$cell_labels))
```

---

## Build spatial neighbour graph

`buildSpatialGraph()` links each cell to all cells within 15 µm (matching the steinbock `--dmax 15` parameter). This replicates the steinbock neighbour graph from coordinates stored in the SPE, avoiding the file-count mismatch in `neighbors/`.

```{r build-graph}
spe <- buildSpatialGraph(
  spe,
  img_id    = "sample_id",
  type      = "expansion",
  threshold = 15,        # µm, matches steinbock --dmax 15
  coords    = c("Pos_X", "Pos_Y"),
  BPPARAM   = BiocParallel::SerialParam()
)

# confirm graph was added
colPairNames(spe)
```


```{r}
spatialCoords(spe)      # should show centroid columns
spatialCoordsNames(spe) # column names
```

---

## Cell-type interaction counts

`countInteractions()` counts, per image, how often each pair of cell types are direct neighbours.
`testInteractions()` runs a permutation test (label-shuffling within each image) to identify pairs that co-occur more or less than expected by chance.

```{r count-interactions}
out_count <- countInteractions(
  spe,
  group_by      = "sample_id",
  label         = "cell_labels",
  colPairName   = "neighborhood",
  method        = "classic"
)

head(out_count)
```

```{r test-interactions}
out_test <- testInteractions(
  spe,
  group_by      = "sample_id",
  label         = "cell_labels",
  colPairName   = "neighborhood",
  method        = "classic",
  iter          = 200,    # permutations per image; raise to 1000 for final run
  BPPARAM       = BiocParallel::SerialParam()
)

head(out_test)
```

---

## Interaction heatmap

Mean interaction score across images. Positive = enriched co-localisation; negative = spatial avoidance.

```{r interaction-heatmap, fig.width=10, fig.height=9}
# testInteractions() returns a Bioconductor DataFrame; convert before dplyr
out_test_df <- as.data.frame(out_test)

# aggregate mean z-score per pair across images (zscore = numeric enrichment score)
agg <- out_test_df |>
  group_by(from_label, to_label) |>
  summarise(mean_zscore = mean(zscore, na.rm = TRUE), .groups = "drop")

# pivot to matrix
mat <- agg |>
  pivot_wider(names_from = to_label, values_from = mean_zscore, values_fill = 0) |>
  tibble::column_to_rownames("from_label") |>
  as.matrix()

# symmetrise into square matrix covering all cell types
all_types <- union(rownames(mat), colnames(mat))
mat_sq <- matrix(0, nrow = length(all_types), ncol = length(all_types),
                 dimnames = list(all_types, all_types))
mat_sq[rownames(mat), colnames(mat)] <- mat

z_lim   <- max(abs(mat_sq), na.rm = TRUE)
col_fun <- colorRamp2(c(-z_lim, 0, z_lim), c("#2166AC", "white", "#D6604D"))
```

```{r}
hm_interactions <- Heatmap(
  mat_sq,
  name            = "Interaction\nscore",
  col             = col_fun,
  cluster_rows    = TRUE,
  cluster_columns = TRUE,
  row_names_gp    = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 9),
  column_title    = "Cell-type spatial interaction scores (mean across images)"
)
png(file.path(plots_dir, "s01_interaction_heatmap.png"),
    width = 10, height = 9, units = "in", res = 150)
draw(hm_interactions)
dev.off()
hm_interactions
```

---

## CD8 — Tumour proximity

Focused look at whether CD8+ T cells are closer to Tumour cells than expected by chance, image by image.

```{r cd8-tumor-proximity, fig.width=7, fig.height=4}
cd8_tumor <- out_test_df |>
  filter((from_label == "CD8" & to_label == "Tumor") |
         (from_label == "Tumor" & to_label == "CD8"))

p_cd8_tumor <- ggplot(cd8_tumor, aes(x = zscore)) +
  geom_histogram(bins = 20, fill = "#2166AC", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  facet_wrap(~paste(from_label, "→", to_label)) +
  labs(
    title = "CD8 ↔ Tumour interaction z-score across images",
    x     = "Z-score (positive = enriched co-localisation)",
    y     = "Number of images"
  ) +
  theme_bw()
p_cd8_tumor
ggsave(file.path(plots_dir, "s02_cd8_tumor_proximity.png"), p_cd8_tumor,
       width = 7, height = 4, dpi = 150)
```

---

## Neighbourhood composition per cell

For each cell, compute the fraction of each cell type among its direct neighbours. These become additional "spatial" features that can be compared to marker-based features.

```{r neighbourhood-composition}
# retrieve the graph edges; colPair() (singular) returns a SelfHits object
# S4Vectors::from()/to() are the stable accessors (avoid @from/@to slot access)
graph    <- colPair(spe, type = "neighborhood")
from_idx <- S4Vectors::from(graph)
to_idx   <- S4Vectors::to(graph)

# cell types of the target cells
to_labels <- spe$cell_labels[to_idx]

# use sample_id + ObjectNumber as the unique cell key (no cell_id column in colData)
cd <- as.data.frame(colData(spe))
cd$cell_key <- paste0(cd$sample_id, "_", cd$ObjectNumber)

from_keys <- cd$cell_key[from_idx]

# count neighbour cell types per source cell
neigh_df <- data.frame(
  cell_key   = from_keys,
  neigh_type = to_labels,
  stringsAsFactors = FALSE
)

# fraction of each type per cell
neigh_comp <- neigh_df |>
  group_by(cell_key, neigh_type) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(cell_key) |>
  mutate(frac = n / sum(n)) |>
  ungroup() |>
  select(cell_key, neigh_type, frac) |>
  pivot_wider(names_from = neigh_type, values_from = frac,
              values_fill = 0, names_prefix = "neigh_")

# attach to colData
cd <- left_join(cd, neigh_comp, by = "cell_key")
```

---

## Do same-marker cells with different neighbours get different labels?

This is the core spatial XAI question. We take cells that are similar in marker expression (same predicted cell type from marker-only model) and ask whether their neighbourhood composition differs by ground-truth label.

```{r marker-vs-spatial, fig.width=10, fig.height=5}
# use exprs assay (compensated, asinh-transformed)
expr_mat <- t(assay(spe, "exprs"))

# exclude DNA/Histone from marker features (same as XGBoost training)
exclude_markers <- c("DNA1", "DNA2", "HistoneH3")
keep_markers    <- !rownames(spe) %in% exclude_markers
expr_mat        <- expr_mat[, keep_markers]

# join with neighbourhood composition
xai_df <- cbind(
  as.data.frame(expr_mat),
  cd[match(paste0(spe$sample_id, "_", spe$ObjectNumber), cd$cell_key), grep("^neigh_", names(cd))]
)
xai_df$cell_labels <- spe$cell_labels

# for each cell type: how much does neighbourhood composition vary?
neigh_cols <- grep("^neigh_", names(xai_df), value = TRUE)

neigh_var <- xai_df |>
  group_by(cell_labels) |>
  summarise(
    across(all_of(neigh_cols),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd   = ~sd(.x, na.rm = TRUE))),
    .groups = "drop"
  )
```

```{r}
# plot mean neighbourhood fraction of Tumour neighbours per cell type
if ("neigh_Tumor" %in% names(xai_df)) {
  p_tumor_neigh <- ggplot(xai_df, aes(x = reorder(cell_labels, neigh_Tumor, median),
                     y = neigh_Tumor)) +
    geom_boxplot(outlier.size = 0.3, fill = "#E8E8E8") +
    labs(
      title = "Fraction of Tumour neighbours by cell type",
      x     = "Cell type",
      y     = "Fraction of direct neighbours that are Tumour"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  p_tumor_neigh
  ggsave(file.path(plots_dir, "s03_tumor_neighbour_fraction.png"), p_tumor_neigh,
         width = 10, height = 5, dpi = 150)
}
```

---

## Spatial feature importance — augmented XGBoost

Train a second XGBoost model that includes neighbourhood composition features and compare feature importance to the marker-only model. This directly answers: "does spatial context add information beyond markers?"

```{r load-classifier}
meta               <- readRDS(file.path(base_path, "ROut", "classifier_xgboost_meta.rds"))
model_markers_only <- readRDS(file.path(base_path, "ROut", "classifier_xgboost.rds"))
```

```{r augmented-model, fig.width=9, fig.height=6}
# build augmented feature matrix: markers + neighbour fractions
spe_keys  <- paste0(spe$sample_id, "_", spe$ObjectNumber)
neigh_mat <- as.matrix(cd[match(spe_keys, cd$cell_key), neigh_cols])
neigh_mat[is.na(neigh_mat)] <- 0

X_markers   <- t(assay(spe, "exprs"))[, meta$feature_names]
X_augmented <- cbind(X_markers, neigh_mat)

y <- as.integer(factor(spe$cell_labels, levels = meta$label_mapping)) - 1L

dtrain_aug <- xgb.DMatrix(data = X_augmented, label = y)

params_aug <- list(
  objective        = "multi:softprob",
  num_class        = length(meta$label_mapping),
  eta              = meta$params$eta,
  max_depth        = meta$params$max_depth,
  subsample        = meta$params$subsample,
  colsample_bytree = meta$params$colsample_bytree,
  eval_metric      = "mlogloss"
)

set.seed(42)
model_aug <- xgb.train(
  params  = params_aug,
  data    = dtrain_aug,
  nrounds = meta$nrounds,
  verbose = 0
)
```

```{r augmented-model, fig.width=9, fig.height=6}

# feature importance: augmented model
imp_aug <- xgb.importance(model = model_aug)

# split into marker vs spatial features
imp_aug$feature_type <- ifelse(
  grepl("^neigh_", imp_aug$Feature), "Spatial (neighbour)", "Marker"
)

p_aug_imp <- ggplot(imp_aug |> slice_max(Gain, n = 30),
       aes(x = reorder(Feature, Gain), y = Gain, fill = feature_type)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Marker" = "#4393C3", "Spatial (neighbour)" = "#D6604D")) +
  labs(
    title = "Top 30 features — augmented model (markers + neighbourhood)",
    x     = NULL,
    y     = "Gain",
    fill  = "Feature type"
  ) +
  theme_bw()
p_aug_imp
ggsave(file.path(plots_dir, "s04_augmented_feature_importance.png"), p_aug_imp,
       width = 9, height = 6, dpi = 150)
```

---

## Test set evaluation: marker-only vs augmented

The augmented model trained above used all 13 spatial images. Here we retrain it on the train-only subset so the held-out test images were never seen during augmented model training — giving a fair comparison against the marker-only model.


```{r}
names(meta)           # should show test_image_ids in the list
meta$test_image_ids   # should show 5 image name strings
unique(as.character(spe$sample_id))
```

```{r test-evaluation}
library(caret)

# Identify which spatial images are held-out test images
test_imgs_spatial  <- intersect(meta$test_image_ids, unique(spe$sample_id))
train_imgs_spatial <- setdiff(unique(spe$sample_id), test_imgs_spatial)

cat(sprintf("Spatial images: %d total — %d train, %d test\n",
            length(unique(spe$sample_id)),
            length(train_imgs_spatial),
            length(test_imgs_spatial)))
```

```{r test-evaluation-run}
if (length(test_imgs_spatial) == 0) {
  warning("No held-out test images found in spatial set.")
} else {
  # Some cell labels in the spatial set may not appear in the training label mapping
  # (cell types exclusive to test images). Exclude those cells from evaluation.
  valid_label <- spe$cell_labels %in% meta$label_mapping
  if (any(!valid_label)) {
    cat(sprintf("Excluding %d cells with labels absent from training mapping: %s\n",
                sum(!valid_label),
                paste(unique(spe$cell_labels[!valid_label]), collapse = ", ")))
  }

  is_train <- as.character(spe$sample_id) %in% train_imgs_spatial & valid_label
  is_test  <- as.character(spe$sample_id) %in% test_imgs_spatial  & valid_label
  cat(sprintf("Train cells: %d | Test cells: %d\n", sum(is_train), sum(is_test)))

  y_train <- as.integer(factor(spe$cell_labels[is_train], levels = meta$label_mapping)) - 1L

  # Retrain augmented model on train images only (reuses X_augmented, X_markers from previous chunk)
  dtrain_aug_eval <- xgb.DMatrix(data = X_augmented[is_train, ], label = y_train)
  set.seed(42)
  model_aug_eval <- xgb.train(
    params  = params_aug,
    data    = dtrain_aug_eval,
    nrounds = meta$nrounds,
    verbose = 0
  )

  true_labels <- factor(spe$cell_labels[is_test], levels = meta$label_mapping)

  # Marker-only model predictions on test set
  dtest_m      <- xgb.DMatrix(data = X_markers[is_test, ])
  pred_m_raw   <- predict(model_markers_only, newdata = dtest_m)
  pred_m_mat   <- matrix(pred_m_raw, ncol = length(meta$label_mapping), byrow = TRUE)
  pred_markers <- factor(meta$label_mapping[max.col(pred_m_mat)], levels = meta$label_mapping)

  # Augmented model predictions on test set
  dtest_a       <- xgb.DMatrix(data = X_augmented[is_test, ])
  pred_a_raw    <- predict(model_aug_eval, newdata = dtest_a)
  pred_a_mat    <- matrix(pred_a_raw, ncol = length(meta$label_mapping), byrow = TRUE)
  pred_aug_test <- factor(meta$label_mapping[max.col(pred_a_mat)], levels = meta$label_mapping)

  cm_markers_test <- caret::confusionMatrix(data = pred_markers,  reference = true_labels)
  cm_aug_test     <- caret::confusionMatrix(data = pred_aug_test, reference = true_labels)

  cat("\n=== Marker-only model — held-out test set ===\n")
  print(cm_markers_test$overall)
  cat("\n")
  print(cm_markers_test$table)

  cat("\n=== Augmented model (markers + neighbourhood) — held-out test set ===\n")
  print(cm_aug_test$overall)
  cat("\n")
  print(cm_aug_test$table)

  cat(sprintf("\nAccuracy delta = %+.5f  |  Kappa delta = %+.5f\n",
              cm_aug_test$overall["Accuracy"] - cm_markers_test$overall["Accuracy"],
              cm_aug_test$overall["Kappa"]    - cm_markers_test$overall["Kappa"]))
  cat("(positive = augmented model better)\n")
}


#              Reference
# Prediction      B BnT CD4 CD8  DC Fibro Igg Mac MacCD204 MacCD209 MDSC mregDC
#   B             0   0   6   7   1     3   1   0        7        0    0      0
#   BnT           0   0   5   7   1     8   0   1        9        1    0      0
#   CD4           0   0   2   7   0     2   0   0        7        1    0      0
#   CD8           0   0   7   7   0     5   0   0       13        0    0      0
#   DC            1   0   6   3   0     3   0   2        6        0    0      0
#   Fibro         0   1   1   2   0     5   0   0        5        0    0      0
#   Igg           2   0  11   9   0     1   0   0        9        1    0      0
#   Mac           0   0   4   3   0     2   0   0        8        1    0      0
#   MacCD204      0   1   5   3   0     4   0   1        7        0    0      0
#   MacCD209      0   0   7   5   3     2   0   1       13        1    0      0
#   MDSC          1   0   6   8   0     8   0   3       11        2    0      0
#   mregDC        1   0   6   5   2     3   0   0        3        5    0      0
#   Neutrophil    0   0  11   3   0     7   0   1        9        1    0      0
#   NK            1   0   8   8   0     2   0   0        3        2    1      0
#   PMN_MDSC      0   1   3   3   0     3   0   0        5        1    0      0
#   SMA           0   0   8   6   0     3   0   2        6        2    0      0
#   Treg          1   0   3   7   0     2   0   1        9        1    0      0
#   Tumor         0   0   9   6   0     4   0   0        7        0    0      0
#   undefined     1   0  10   2   0     4   0   1        3        1    1      0
#   Vasculature   0   0   2   4   0     6   0   0        9        1    0      0
#   vCAF          0   0   4   8   0     1   0   0       10        0    0      0
#              Reference
# Prediction    Neutrophil  NK PMN_MDSC SMA Treg Tumor undefined Vasculature vCAF
#   B                    0   0        2   1    0   293         4           1    4
#   BnT                  0   0        1   3    1   267         2           3    4
#   CD4                  0   0        0   1    1   232         3           1    6
#   CD8                  1   0        1   3    3   300        10           3    5
#   DC                   0   0        1   2    3   266         3           1    2
#   Fibro                0   0        1   3    1   232         3           2    4
#   Igg                  0   0        0   3    1   279         1           3    4
#   Mac                  0   0        2   1    1   254         2           3    0
#   MacCD204             0   0        0   4    1   220         2           3    5
#   MacCD209             0   0        0   2    1   290         1           0    5
#   MDSC                 0   0        0   3    2   269         5           1    1
#   mregDC               0   0        0   3    3   222         3           0    8
#   Neutrophil           0   0        1   1    1   288         5           2    0
#   NK                   0   0        0   2    2   281         1           1    5
#   PMN_MDSC             0   0        0   0    5   229         4           1    4
#   SMA                  0   0        2   1    3   276         4           2    5
#   Treg                 0   0        1   1    1   289         4           1    4
#   Tumor                1   0        0   1    1   235         4           4    4
#   undefined            0   0        2   6    2   292         3           1    1
#   Vasculature          0   0        1   2    2   277         3           0    2
#   vCAF                 1   0        2   2    2   234         3           2    2

# Accuracy delta = +0.00016  |  Kappa delta = -0.00144
```

---

## Summary

```{r summary}
cat("=== Spatial XAI Summary ===\n\n")

cat("Interaction test: cell-type pairs with significant co-localisation:\n")
sig_pairs <- out_test_df |>
  group_by(from_label, to_label) |>
  summarise(
    mean_zscore      = mean(zscore, na.rm = TRUE),
    frac_enriched    = mean(interaction == TRUE, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(abs(mean_zscore) > 1) |>
  arrange(desc(abs(mean_zscore)))

print(sig_pairs)
```

```{r summary}
cat("\nSpatial feature contribution in augmented model:\n")
spatial_gain <- imp_aug |>
  group_by(feature_type) |>
  summarise(total_gain = sum(Gain), .groups = "drop") |>
  mutate(pct = round(100 * total_gain / sum(total_gain), 1))
print(spatial_gain)
```
