---
title: "TME Heterogeneity & Anomaly Detection"
author: "Oleksandra Tsepilova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview
This script pivots to mapping intratumoral heterogeneity across the cohort. Because every image comes from a tumor-bearing patient, we apply a Centered Log-Ratio (CLR) transformation to mathematically neutralize the dominant tumor mass and reveal the underlying variance in the immune and stromal compartments. We then run a Principal Component Analysis (PCA) to flag biologically extreme Tumor Microenvironment (TME) phenotypes for team handoff.

### 1. Load Libraries and the QC-Passed Tumor Dataset

```{r load-data, message=FALSE}
library(SingleCellExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)

base_path <- file.path(
  path.expand("."), 
  "data", 
  "Panel_2_10"
)

# Setup output directories
plot_dir <- file.path(base_path, "ROut", "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

# Load the mathematically verified, QC-passed dataset
# Note: "undefined" machine-error cells and blank slides were purged
spe_clean <- readRDS(file.path(base_path, "ROut", "spe_master_classified.rds"))

# Double-check factor levels are clean
spe_clean$final_label <- droplevels(spe_clean$final_label)

message("Loaded QC-passed tumor dataset. Total cells ready for PCA: ", ncol(spe_clean))
```

### 2. CLR Transformation and PCA

We convert the raw cell counts into compositional proportions. Because this data is dominated by tumor cells, we apply a CLR transformation to center the data around the geometric mean of each image, ensuring rare immune populations drive the PCA variance appropriately.

```{r clr-pca}
# Generate the raw count matrix (Images as rows, Cell Types as columns)
df_counts <- as.data.frame(colData(spe_clean)) %>%
  group_by(sample_id, final_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = final_label, values_from = Count, values_fill = 0)

# Extract matrix and add a pseudocount of 1 to avoid log(0) errors
matrix_counts <- as.matrix(df_counts[, -1])
rownames(matrix_counts) <- df_counts$sample_id
matrix_pseudo <- matrix_counts + 1

# Convert to relative proportions per image
matrix_prop <- sweep(matrix_pseudo, 1, rowSums(matrix_pseudo), "/")

# Apply Centered Log-Ratio (CLR) Transformation
clr_matrix <- t(apply(matrix_prop, 1, function(x) {
  log(x / exp(mean(log(x))))
}))

# Run Principal Component Analysis
pca_res <- prcomp(clr_matrix, center = TRUE, scale. = FALSE)

# Extract variance explained for the axes
var_explained <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

# Format the results into a plotting dataframe
pca_df <- as.data.frame(pca_res$x)
pca_df$sample_id <- rownames(pca_df)
```

### 3. Isolate Outliers (Biological Extremes)

We define an anomaly as any tumor image that falls more than 2 standard deviations away from the cohort mean on either Principal Component 1 (the axis of maximum variance) or Principal Component 2. These represent extreme TME phenotypes (e.g., hyper-inflamed or completely immune-deserted tissues).

```{r isolate-outliers}
# Calculate statistical boundaries
sd_pc1 <- sd(pca_df$PC1)
sd_pc2 <- sd(pca_df$PC2)
mean_pc1 <- mean(pca_df$PC1)
mean_pc2 <- mean(pca_df$PC2)

# Flag the anomalies
pca_df <- pca_df %>%
  mutate(
    is_anomaly = abs(PC1 - mean_pc1) > (2 * sd_pc1) | abs(PC2 - mean_pc2) > (2 * sd_pc2)
  )

anomalous_images <- pca_df$sample_id[pca_df$is_anomaly]

message("Anomaly Detection Complete.")
message("Identified ", length(anomalous_images), " extreme TME phenotypes out of ", nrow(pca_df), " total QC-passed tumor images.")
```

### 4. Visualize the Anomaly Map

We plot PC1 vs PC2 and visually highlight the 2-Sigma baseline boundary. Any points falling outside this blue box represent patient tissues with highly unusual biology requiring deeper investigation.

```{r plot-pca-anomalies}
pdf(file.path(plot_dir, "09_PCA_Anomaly_Map.pdf"), width = 12, height = 8)
ggplot(pca_df, aes(x = PC1, y = PC2, fill = is_anomaly)) +
  # Draw boundary boxes representing the 2-Sigma thresholds
  annotate("rect", 
           xmin = mean_pc1 - (2*sd_pc1), xmax = mean_pc1 + (2*sd_pc1), 
           ymin = mean_pc2 - (2*sd_pc2), ymax = mean_pc2 + (2*sd_pc2), 
           alpha = 0.1, fill = "blue") +
  geom_point(shape = 21, size = 4, color = "black", alpha = 0.8) +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  # Only label the anomalous points to keep the plot clean
  geom_text_repel(
    data = subset(pca_df, is_anomaly),
    aes(label = sample_id),
    size = 3,
    box.padding = 0.5,
    max.overlaps = Inf
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "Intratumoral Heterogeneity & Anomaly Detection (CLR-PCA)",
    subtitle = paste0("Red points indicate > 2 SD from the mean. N = ", length(anomalous_images), " extreme phenotypes flagged for handoff."),
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    fill = "Extreme Phenotype"
  ) +
  theme(legend.position = "bottom")
dev.off()

message("PCA Anomaly Map saved to ROut/plots.")
```

### 5. Generate the Team Handoff Manifest

We isolate the `sample_id`s of the anomalous images and export them to a dedicated text file containing instructions for Person 2 and Person 3.

```{r team-handoff}
handoff_file <- file.path(base_path, "ROut", "Team_Handoff_Anomalies.txt")

# Write the handoff directives
writeLines(c(
  "==================================================",
  "TME ANOMALY HANDOFF MANIFEST",
  paste("Generated on:", Sys.time()),
  "==================================================",
  "",
  "The following tumor images exhibited extreme cellular compositions",
  "following CLR-PCA dimensionality reduction, representing significant",
  "deviations from the baseline cohort phenotype.",
  "",
  "--> ACTION REQUIRED - Inna (Uncertainty):",
  "Please subset these images and verify if the XGBoost softmax probabilities",
  "exhibited unusual uncertainty distributions in these specific tissues.",
  "Check if the model was forced to extrapolate on out-of-distribution biological data.",
  "",
  "--> ACTION REQUIRED - Anna (Spatial Context):",
  "Please map the physical neighborhood graphs (KNN) for these specific outliers",
  "to determine if this represents a biologically unique Tumor Microenvironment",
  "(e.g., hyper-inflamed state, invasive margin) or a physical slide artifact.",
  "",
  "ANOMALOUS SAMPLE IDs:",
  "--------------------------------------------------"
), con = handoff_file)

# Append the actual IDs
cat(paste(anomalous_images, collapse = "\n"), file = handoff_file, append = TRUE)

message("Handoff manifest successfully written to: ", handoff_file)
```