---
title: "Phase 3: RAW Proportions, QC Insights & CLEAN Phenotype Clustering"
author: "Oleksandra Tsepilova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview
This script unifies datasets to generate a "RAW" proportion plot and a Quality Control (QC) marker heatmap, explicitly proving why corrupted slides were flagged as "undefined". Then, it aggressively cleans the dataset and applies a Centered Log-Ratio (CLR) transformation to cluster the remaining healthy tissues into distinct biological phenotypes.

### 1. Load Libraries, Data, and Palette

```{r load-data, message=FALSE}
library(SingleCellExperiment)
library(dplyr)
library(purrr)
library(ggplot2)
library(tidyr)
library(ComplexHeatmap)
library(circlize)

base_path <- file.path(
  path.expand("~"), 
  "Downloads", 
  "XGBoost_classifier_updated", 
  "Panel_2_10"
)

plot_dir <- file.path(base_path, "ROut", "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

cell_type_colors <- c(
  B = "red3", BnT = "indianred4", CD4 = "palevioletred1", CD8 = "darkorchid",
  Treg = "maroon4", NK = "yellow", Igg = "darkred", DC = "green4",
  Mac = "darkolivegreen", MacCD209 = "olivedrab1", MacCD204 = "olivedrab3",
  mregDC = "green", Neutrophil = "blue1", PMN_MDSC = "deepskyblue",
  MDSC = "aquamarine4", Vasculature = "brown2", vCAF = "tomato",
  Tumor = "sienna4", SMA = "goldenrod2", Fibro = "chocolate1",
  unlabelled = "gray", undefined = "black"
)

official_labels <- c(
  "B", "BnT", "CD4", "CD8", "DC", "Fibro", "Igg", "Mac", "MacCD204", 
  "MacCD209", "MDSC", "mregDC", "Neutrophil", "NK", "PMN_MDSC", "SMA", 
  "Treg", "Tumor", "undefined", "Vasculature", "vCAF"
)
```

### 2. Extract and Sanitize Ground Truth Labels

```{r sanitize-gates}
gates <- list.files(file.path(base_path, "gates"), pattern = "\\.rds$", recursive = TRUE, full.names = TRUE)

gate_labels <- purrr::map_dfr(gates, function(path) {
  x <- readRDS(path)
  data.frame(
    sample_id = as.character(x$sample_id),
    ObjectNumber = as.character(x$ObjectNumber),
    cell_labels = as.character(x$cytomapper_CellLabel),
    row.names = NULL
  )
})

gate_labels$cell_labels <- trimws(gate_labels$cell_labels)
gate_labels$cell_labels <- sub("_[0-9]+[a-zA-Z]?$", "", gate_labels$cell_labels)
gate_labels$cell_labels <- sub(" \\([0-9]+\\)$", "", gate_labels$cell_labels)
gate_labels$cell_labels[gate_labels$cell_labels == "Cd4"] <- "CD4"
gate_labels$cell_labels[gate_labels$cell_labels == "Cd8"] <- "CD8"
gate_labels$cell_labels[gate_labels$cell_labels == "MDSC_"] <- "MDSC"
gate_labels$cell_labels[!(gate_labels$cell_labels %in% official_labels)] <- "undefined"
gate_labels$cell_id <- paste0(gate_labels$sample_id, "_", gate_labels$ObjectNumber)
```

### 3. Master Dataset Consolidation (Keeping RAW Data)

```{r consolidate-datasets}
spe_base <- readRDS(file.path(base_path, "ROut", "spe.rds"))
spe_pred <- readRDS(file.path(base_path, "ROut", "spe_unseen_predicted.rds"))

spe_base$cell_id <- paste0(spe_base$sample_id, "_", spe_base$ObjectNumber)
colnames(spe_base) <- spe_base$cell_id

spe_base$final_label <- "undefined"
spe_base$label_source <- "Unknown"

match_pred <- match(spe_base$cell_id, spe_pred$cell_id)
has_pred <- !is.na(match_pred)
spe_base$final_label[has_pred] <- as.character(spe_pred$predicted_label[match_pred[has_pred]])
spe_base$label_source[has_pred] <- "XGBoost_Prediction"

match_gt <- match(spe_base$cell_id, gate_labels$cell_id)
has_gt <- !is.na(match_gt)
spe_base$final_label[has_gt] <- gate_labels$cell_labels[match_gt[has_gt]]
spe_base$label_source[has_gt] <- "Manual_GroundTruth"

spe_base$final_label <- factor(spe_base$final_label, levels = official_labels)
```

### 4. INSIGHT 1: Plotting the RAW Proportions

```{r plot-raw-insight}
df_raw <- as.data.frame(colData(spe_base))

prop_df_raw <- df_raw %>%
  group_by(sample_id, final_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(sample_id) %>%
  mutate(Proportion = Count / sum(Count))

pdf(file.path(plot_dir, "07a_RAW_Proportions_Per_Image.pdf"), width = 16, height = 10)
ggplot(prop_df_raw, aes(x = sample_id, y = Proportion, fill = final_label)) +
  geom_bar(stat = "identity", width = 0.9) +
  scale_fill_manual(values = cell_type_colors) +
  theme_classic(base_size = 12) +
  labs(
    title = "RAW Cell Type Proportions (Including Model QC Flags)",
    subtitle = "Solid black bars indicate corrupted slides safely caught by the model as 'undefined'.",
    x = "Image ID",
    y = "Relative Proportion"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    legend.position = "right",
    legend.title = element_blank()
  )
dev.off()
```

### 5. INSIGHT 2: Corrupted Slides Marker Expression Heatmap

We extract the mean protein expression for the "dead" images and compare them against a baseline average of all healthy images to prove visual signal loss.

```{r qc-marker-heatmap}
# Identify dead/blank images (>90% undefined)
undefined_stats <- df_raw %>%
  group_by(sample_id) %>%
  summarise(pct_undefined = sum(final_label == "undefined") / n())

dead_images <- undefined_stats$sample_id[undefined_stats$pct_undefined > 0.90]

if(length(dead_images) > 0) {
  
  # Extract raw marker matrix and combine with sample IDs
  exprs_matrix <- t(assay(spe_base, "exprs"))
  exprs_df <- as.data.frame(exprs_matrix)
  exprs_df$sample_id <- spe_base$sample_id
  
  # Calculate mean expression for the dead images
  dead_exprs <- exprs_df %>%
    filter(sample_id %in% dead_images) %>%
    group_by(sample_id) %>%
    summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
  
  # Calculate a single "Healthy Baseline" average
  healthy_baseline <- exprs_df %>%
    filter(!(sample_id %in% dead_images)) %>%
    summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE))) %>%
    mutate(sample_id = "HEALTHY_AVERAGE_BASELINE")
  
  # Combine into a plotting matrix
  qc_matrix_df <- bind_rows(dead_exprs, healthy_baseline)
  rownames(qc_matrix_df) <- qc_matrix_df$sample_id
  qc_matrix_df$sample_id <- NULL
  qc_matrix <- as.matrix(qc_matrix_df)
  
  # Plot the QC Heatmap
  pdf(file.path(plot_dir, "07b_QC_Dead_Images_Marker_Heatmap.pdf"), width = 12, height = 5)
  ht <- Heatmap(
    qc_matrix,
    name = "Mean\nExpression",
    col = colorRamp2(c(0, 1, 3), c("#2166ac", "white", "#b2182b")), # Blue (0) to Red (High)
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    row_title = "Corrupted Slides",
    row_title_gp = gpar(fontface = "bold", fontsize = 12),
    column_title = "Protein Markers (Proving Signal Flatline)",
    column_title_gp = gpar(fontface = "bold", fontsize = 12),
    rect_gp = gpar(col = "grey90", lwd = 0.5),
    row_names_side = "left"
  )
  draw(ht)
  dev.off()
  
  message("QC Marker Heatmap saved! Flatline signal proven.")
}
```

### 6. Data Cleaning for CLR Math

We purge the dead images and the "undefined" cells to protect the clustering algorithms.

```{r purge-and-clean}
# Subset the SPE to drop dead images
if(length(dead_images) > 0) {
  spe_master_clean <- spe_base[, !(spe_base$sample_id %in% dead_images)]
} else {
  spe_master_clean <- spe_base
}

# CRITICAL: Drop remaining 'undefined' cells to protect the math
spe_master_clean <- spe_master_clean[, spe_master_clean$final_label != "undefined"]
spe_master_clean$final_label <- droplevels(spe_master_clean$final_label)

saveRDS(spe_master_clean, file.path(base_path, "ROut", "spe_master_classified.rds"))
message("Data successfully cleaned! Ready for CLR.")
```

### 7. CLEAN TME Phenotype Clustering (CLR Transformation)

```{r clr-heatmap}
df_master_clean <- as.data.frame(colData(spe_master_clean))

# Prepare Count Matrix
df_counts <- df_master_clean %>%
  group_by(sample_id, final_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = final_label, values_from = Count, values_fill = 0)

matrix_counts <- as.matrix(df_counts[, -1])
rownames(matrix_counts) <- df_counts$sample_id

# Add Pseudocount and convert to proportions
matrix_pseudo <- matrix_counts + 1
matrix_prop <- sweep(matrix_pseudo, 1, rowSums(matrix_pseudo), "/")

# Apply CLR Transformation
clr_matrix <- t(apply(matrix_prop, 1, function(x) {
  log(x / exp(mean(log(x))))
}))

# Generate ComplexHeatmap
clr_min <- floor(min(clr_matrix))
clr_max <- ceiling(max(clr_matrix))
col_fun <- colorRamp2(c(clr_min, 0, clr_max), c("#2166ac", "white", "#b2182b"))

pdf(file.path(plot_dir, "08_CLEAN_CLR_Phenotype_Heatmap.pdf"), width = 14, height = 10)
Heatmap(
  clr_matrix,
  name = "CLR\nEnrichment",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 10),
  row_title = "Patient Tissue Images\n(Clustered by TME Phenotype)",
  row_title_gp = gpar(fontsize = 12, fontface = "bold"),
  column_title = "Cell Types",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5)
)
dev.off()

message("Complete! Raw bins, QC Heatmap, and Clean CLR Heatmap saved.")
```