---
title: "RAW Proportions & CLEAN Phenotype Clustering"
author: "Oleksandra Tsepilova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview
This script unifies the datasets to first generate a "RAW" proportion plot, highlighting the corrupted slides flagged by the model as "undefined". Then, it aggressively cleans the dataset and applies a Centered Log-Ratio (CLR) transformation to cluster the remaining healthy tissues into distinct biological phenotypes.

### 1. Load Libraries, Data, and Palette

```{r load-data, message=FALSE}
library(SingleCellExperiment)
library(dplyr)
library(purrr)
library(ggplot2)
library(tidyr)
library(ComplexHeatmap)
library(circlize)

base_path <- file.path(
  path.expand("."), 
  "data", 
  "Panel_2_10"
)

plot_dir <- file.path(base_path, "ROut", "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

cell_type_colors <- c(
  B = "red3", BnT = "indianred4", CD4 = "palevioletred1", CD8 = "darkorchid",
  Treg = "maroon4", NK = "yellow", Igg = "darkred", DC = "green4",
  Mac = "darkolivegreen", MacCD209 = "olivedrab1", MacCD204 = "olivedrab3",
  mregDC = "green", Neutrophil = "blue1", PMN_MDSC = "deepskyblue",
  MDSC = "aquamarine4", Vasculature = "brown2", vCAF = "tomato",
  Tumor = "sienna4", SMA = "goldenrod2", Fibro = "chocolate1",
  unlabelled = "gray", undefined = "black"
)

official_labels <- c(
  "B", "BnT", "CD4", "CD8", "DC", "Fibro", "Igg", "Mac", "MacCD204", 
  "MacCD209", "MDSC", "mregDC", "Neutrophil", "NK", "PMN_MDSC", "SMA", 
  "Treg", "Tumor", "undefined", "Vasculature", "vCAF"
)
```

### 2. Extract and Sanitize Ground Truth Labels

```{r sanitize-gates}
gates <- list.files(file.path(base_path, "gates"), pattern = "\\.rds$", recursive = TRUE, full.names = TRUE)

gate_labels <- purrr::map_dfr(gates, function(path) {
  x <- readRDS(path)
  data.frame(
    sample_id = as.character(x$sample_id),
    ObjectNumber = as.character(x$ObjectNumber),
    cell_labels = as.character(x$cytomapper_CellLabel),
    row.names = NULL
  )
})

gate_labels$cell_labels <- trimws(gate_labels$cell_labels)
gate_labels$cell_labels <- sub("_[0-9]+[a-zA-Z]?$", "", gate_labels$cell_labels)
gate_labels$cell_labels <- sub(" \\([0-9]+\\)$", "", gate_labels$cell_labels)
gate_labels$cell_labels[gate_labels$cell_labels == "Cd4"] <- "CD4"
gate_labels$cell_labels[gate_labels$cell_labels == "Cd8"] <- "CD8"
gate_labels$cell_labels[gate_labels$cell_labels == "MDSC_"] <- "MDSC"
gate_labels$cell_labels[!(gate_labels$cell_labels %in% official_labels)] <- "undefined"
gate_labels$cell_id <- paste0(gate_labels$sample_id, "_", gate_labels$ObjectNumber)
```

### 3. Master Dataset Consolidation (Keeping RAW Data)

```{r consolidate-datasets}
spe_base <- readRDS(file.path(base_path, "ROut", "spe.rds"))
spe_pred <- readRDS(file.path(base_path, "ROut", "spe_unseen_predicted.rds"))

spe_base$cell_id <- paste0(spe_base$sample_id, "_", spe_base$ObjectNumber)
colnames(spe_base) <- spe_base$cell_id

spe_base$final_label <- "undefined"
spe_base$label_source <- "Unknown"

match_pred <- match(spe_base$cell_id, spe_pred$cell_id)
has_pred <- !is.na(match_pred)
spe_base$final_label[has_pred] <- as.character(spe_pred$predicted_label[match_pred[has_pred]])
spe_base$label_source[has_pred] <- "XGBoost_Prediction"

match_gt <- match(spe_base$cell_id, gate_labels$cell_id)
has_gt <- !is.na(match_gt)
spe_base$final_label[has_gt] <- gate_labels$cell_labels[match_gt[has_gt]]
spe_base$label_source[has_gt] <- "Manual_GroundTruth"

spe_base$final_label <- factor(spe_base$final_label, levels = official_labels)
```

### 4. INSIGHT: Plotting the RAW Proportions

We plot the uncleaned data exactly as it is, explicitly showing how the model flagged the corrupted slides by dumping them into the "undefined" category.

```{r plot-raw-insight}
df_raw <- as.data.frame(colData(spe_base))

prop_df_raw <- df_raw %>%
  group_by(sample_id, final_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(sample_id) %>%
  mutate(Proportion = Count / sum(Count))

pdf(file.path(plot_dir, "07_RAW_Proportions_Per_Image.pdf"), width = 16, height = 10)
ggplot(prop_df_raw, aes(x = sample_id, y = Proportion, fill = final_label)) +
  geom_bar(stat = "identity", width = 0.9) +
  scale_fill_manual(values = cell_type_colors) +
  theme_classic(base_size = 12) +
  labs(
    title = "RAW Cell Type Proportions (Including Model QC Flags)",
    subtitle = "Solid black bars indicate corrupted slides safely caught by the model as 'undefined'.",
    x = "Image ID",
    y = "Relative Proportion"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    legend.position = "right",
    legend.title = element_blank()
  )
dev.off()

message("Raw proportions plot saved!")
```

### 5. Data Cleaning for CLR Math

Now that we have the insight visual, we must purge the dead images and the "undefined" cells. If we don't, the CLR math will treat "undefined" as a biological feature and cluster images based on machine error rather than actual biology.

```{r purge-and-clean}
# 1. Automatically identify dead/blank images (>90% undefined)
undefined_stats <- df_raw %>%
  group_by(sample_id) %>%
  summarise(pct_undefined = sum(final_label == "undefined") / n())

dead_images <- undefined_stats$sample_id[undefined_stats$pct_undefined > 0.90]

# 2. Subset the SPE to drop dead images
if(length(dead_images) > 0) {
  spe_master_clean <- spe_base[, !(spe_base$sample_id %in% dead_images)]
} else {
  spe_master_clean <- spe_base
}

# 3. CRITICAL: Drop remaining 'undefined' cells to protect the math
spe_master_clean <- spe_master_clean[, spe_master_clean$final_label != "undefined"]
spe_master_clean$final_label <- droplevels(spe_master_clean$final_label)

saveRDS(spe_master_clean, file.path(base_path, "ROut", "spe_master_classified.rds"))
message("Data successfully cleaned! Ready for CLR.")
```

### 6. CLEAN TME Phenotype Clustering (CLR Transformation)

We apply a Centered Log-Ratio transformation to the healthy cohort. This prevents massive Tumor/Fibroblast counts from masking subtle fluctuations in critical rare populations, allowing us to accurately cluster the images into biological phenotypes.

```{r clr-heatmap}
df_master_clean <- as.data.frame(colData(spe_master_clean))

# Prepare Count Matrix
df_counts <- df_master_clean %>%
  group_by(sample_id, final_label) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = final_label, values_from = Count, values_fill = 0)

matrix_counts <- as.matrix(df_counts[, -1])
rownames(matrix_counts) <- df_counts$sample_id

# Add Pseudocount and convert to proportions
matrix_pseudo <- matrix_counts + 1
matrix_prop <- sweep(matrix_pseudo, 1, rowSums(matrix_pseudo), "/")

# Apply CLR Transformation
clr_matrix <- t(apply(matrix_prop, 1, function(x) {
  log(x / exp(mean(log(x))))
}))

# Generate ComplexHeatmap
clr_min <- floor(min(clr_matrix))
clr_max <- ceiling(max(clr_matrix))
col_fun <- colorRamp2(c(clr_min, 0, clr_max), c("#2166ac", "white", "#b2182b"))

pdf(file.path(plot_dir, "08_CLEAN_CLR_Phenotype_Heatmap.pdf"), width = 14, height = 10)
Heatmap(
  clr_matrix,
  name = "CLR\nEnrichment",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 10),
  row_title = "Patient Tissue Images\n(Clustered by TME Phenotype)",
  row_title_gp = gpar(fontsize = 12, fontface = "bold"),
  column_title = "Cell Types",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  rect_gp = gpar(col = "grey90", lwd = 0.5)
)
dev.off()

message("Complete! Raw bins and Clean CLR Heatmap saved.")
```