---
title: "Predicting the Unlabelled Cells"
author: "Oleksandra Tsepilova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview
This script loads the entire compensated dataset, programmatically isolates cells without human annotations, and runs the trained XGBoost model. 

### 1. Load Libraries, Data, and Official Labels

```{r load-data}
library(SingleCellExperiment)
library(xgboost)
library(dplyr)
library(purrr)

base_path <- file.path(
  path.expand("."), 
  "data", 
  "Panel_2_10"
)

# Load the full, compensated dataset
spe <- readRDS(file.path(base_path, "ROut", "spe.rds"))

# Create unique cell IDs across the entire dataset
spe$cell_id <- paste0(spe$sample_id, "_", spe$ObjectNumber)
colnames(spe) <- spe$cell_id

# Load the trained model
final_model <- readRDS(file.path(base_path, "classifier_xgboost.rds"))

# Define the EXACT 21 classes derived from the original training matrix
official_labels <- c(
  "B", "BnT", "CD4", "CD8", "DC", "Fibro", "Igg", "Mac", "MacCD204", 
  "MacCD209", "MDSC", "mregDC", "Neutrophil", "NK", "PMN_MDSC", "SMA", 
  "Treg", "Tumor", "undefined", "Vasculature", "vCAF"
)
```

### 2. Identify and Isolate Unlabelled Cells

We exclude any cell that has a manual gate, as well as the blank image identified during QC.

```{r filter-unseen}
# Read the manual gate files to find the cells that were already used
gates <- list.files(
  file.path(base_path, "gates"), 
  pattern = "\\.rds$", 
  full.names = TRUE, 
  recursive = TRUE
)

gate_labels <- purrr::map_dfr(gates, function(path) {
  x <- readRDS(path)
  data.frame(
    sample_id = as.character(x$sample_id),
    ObjectNumber = as.character(x$ObjectNumber),
    row.names = NULL
  )
})

# Reconstruct the cell_ids that have manual annotations
gated_cell_ids <- paste0(gate_labels$sample_id, "_", gate_labels$ObjectNumber)

# Define the known blank image to drop (known from train script)
blank_image <- "10123883-USL-VAR-TIS-01-IMC-02_005"

# Subset the SPE object: Keep cells NOT in the gated list, and NOT in the blank image
unseen_spe <- spe[, !(colnames(spe) %in% gated_cell_ids) & spe$sample_id != blank_image]

message("Total cells in original dataset: ", ncol(spe))
message("Total unlabelled cells to predict: ", ncol(unseen_spe))
```

### 3. Format the Feature Matrix

```{r format-matrix}
# Filter out DNA and Histone markers exactly as done in training
marker_keep <- !grepl("DNA|Histone", rownames(unseen_spe), ignore.case = TRUE)

# Transpose so cells are rows and markers are columns
X_unseen <- t(assay(unseen_spe, "exprs")[marker_keep, ])

# Convert to XGBoost DMatrix format
Xy_unseen <- xgboost::xgb.DMatrix(data = X_unseen)
```

### 4. Execute Predictions with Dynamic Matrix Reshaping

Here we run the predictions by XGBoost. By checking whether the rows or columns sum to 1, we prevent the data from being scrambled.

```{r predict-labels}
message("Running XGBoost mass prediction...")
raw_preds <- predict(final_model, newdata = Xy_unseen)

num_classes <- length(official_labels)

# Test both reshaping directions to find the mathematically valid orientation
prob_byrow <- matrix(raw_preds, ncol = num_classes, byrow = TRUE)
prob_bycol <- matrix(raw_preds, ncol = num_classes, byrow = FALSE)

if (mean(abs(rowSums(prob_byrow) - 1)) < mean(abs(rowSums(prob_bycol) - 1))) {
  prob_matrix <- prob_byrow
} else {
  prob_matrix <- prob_bycol
}

# Safety check: ensure the resulting matrix is structurally sound
stopifnot(mean(abs(rowSums(prob_matrix) - 1)) < 1e-6)

# Assign the class with the highest probability
predicted_labels <- factor(
  official_labels[max.col(prob_matrix, ties.method = "first")],
  levels = official_labels
)

# Extract the maximum probability as the confidence score
prediction_confidence <- apply(prob_matrix, 1, max)
```

### 5. Append Results and Save

```{r save-results}
# Save predictions and confidence back to the metadata
unseen_spe$predicted_label <- predicted_labels
unseen_spe$prediction_confidence <- prediction_confidence

# Save the final predicted object
out_file <- file.path(base_path, "ROut", "spe_unseen_predicted.rds")
saveRDS(unseen_spe, out_file)

message("Predicted SPE saved to: ", out_file)
```