---
title: "Spatial Validation (Training vs. Unseen with Ground Truth)"
author: "Oleksandra Tsepilova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview
This script loads the predicted unseen cells and samples them alongside images from the training cohort.

### 1. Load Libraries, Data, and Setup Output Directories

```{r load-data}
library(SingleCellExperiment)
library(cytomapper)
library(xgboost)
library(dplyr)
library(purrr)

base_path <- file.path(
  path.expand("."), 
  "data", 
  "Panel_2_10"
)

# Create a dedicated directory to immediately save our high-res plots
plot_dir <- file.path(base_path, "ROut", "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

# Load the UNSEEN predicted data
spe_unseen <- readRDS(file.path(base_path, "ROut", "spe_unseen_predicted.rds"))

# Load the original base dataset and model
spe_base <- readRDS(file.path(base_path, "ROut", "spe.rds"))
final_model <- readRDS(file.path(base_path, "classifier_xgboost.rds"))

# Standardized color palette
cell_type_colors <- c(
  B = "red3", BnT = "indianred4", CD4 = "palevioletred1", CD8 = "darkorchid",
  Treg = "maroon4", NK = "yellow", Igg = "darkred", DC = "green4",
  Mac = "darkolivegreen", MacCD209 = "olivedrab1", MacCD204 = "olivedrab3",
  mregDC = "green", Neutrophil = "blue1", PMN_MDSC = "deepskyblue",
  MDSC = "aquamarine4", Vasculature = "brown2", vCAF = "tomato",
  Tumor = "sienna4", SMA = "goldenrod2", Fibro = "chocolate1",
  unlabelled = "gray", undefined = "black"
)

# Define the exact 21 classes derived from the original training matrix
official_labels <- c(
  "B", "BnT", "CD4", "CD8", "DC", "Fibro", "Igg", "Mac", "MacCD204", 
  "MacCD209", "MDSC", "mregDC", "Neutrophil", "NK", "PMN_MDSC", "SMA", 
  "Treg", "Tumor", "undefined", "Vasculature", "vCAF"
)
```

### 2. Sample Unseen and Training Images

```{r prepare-samples}
set.seed(0)

# --- A. Prepare Unseen Images ---
unseen_image_ids <- unique(spe_unseen$sample_id)
sampled_unseen <- sample(unseen_image_ids, 2)
val_unseen <- spe_unseen[, spe_unseen$sample_id %in% sampled_unseen]
val_unseen$image_status <- "Unseen Image"

# --- B. Prepare Training Images and STRICT Ground Truth Labels ---
gates <- list.files(file.path(base_path, "gates"), pattern = "\\.rds$", recursive = TRUE, full.names = TRUE)

gate_labels <- purrr::map_dfr(gates, function(path) {
  x <- readRDS(path)
  data.frame(
    sample_id = as.character(x$sample_id),
    ObjectNumber = as.character(x$ObjectNumber),
    cell_labels = as.character(x$cytomapper_CellLabel),
    row.names = NULL
  )
})

# Aggressively sanitize the manual human annotations
gate_labels$cell_labels <- trimws(gate_labels$cell_labels)
gate_labels$cell_labels <- sub("_[0-9]+[a-zA-Z]?$", "", gate_labels$cell_labels)
gate_labels$cell_labels <- sub(" \\([0-9]+\\)$", "", gate_labels$cell_labels)
gate_labels$cell_labels[gate_labels$cell_labels == "Cd4"] <- "CD4"
gate_labels$cell_labels[gate_labels$cell_labels == "Cd8"] <- "CD8"
gate_labels$cell_labels[gate_labels$cell_labels == "MDSC_"] <- "MDSC"
gate_labels$cell_labels[!(gate_labels$cell_labels %in% official_labels)] <- "undefined"
gate_labels$cell_id <- paste0(gate_labels$sample_id, "_", gate_labels$ObjectNumber)

gated_ids <- unique(gate_labels$sample_id)
valid_training_ids <- intersect(gated_ids, unique(spe_base$sample_id))
training_image_ids <- setdiff(valid_training_ids, "10123883-USL-VAR-TIS-01-IMC-02_005")

sampled_train <- sample(training_image_ids, 2)
val_train <- spe_base[, spe_base$sample_id %in% sampled_train]
val_train$cell_id <- paste0(val_train$sample_id, "_", val_train$ObjectNumber)
colnames(val_train) <- val_train$cell_id

# Attach the sanitized ground truth labels
val_train$cell_labels <- "unlabelled"
match_idx <- match(val_train$cell_id, gate_labels$cell_id)
has_gate <- !is.na(match_idx)
val_train$cell_labels[has_gate] <- gate_labels$cell_labels[match_idx[has_gate]]

# Generate predictions
marker_keep <- !grepl("DNA|Histone", rownames(val_train), ignore.case = TRUE)
X_train <- t(assay(val_train, "exprs")[marker_keep, ])
raw_preds <- predict(final_model, newdata = xgboost::xgb.DMatrix(data = X_train))

# Dynamic matrix reshaping
num_classes <- length(official_labels)
prob_byrow <- matrix(raw_preds, ncol = num_classes, byrow = TRUE)
prob_bycol <- matrix(raw_preds, ncol = num_classes, byrow = FALSE)

if (mean(abs(rowSums(prob_byrow) - 1)) < mean(abs(rowSums(prob_bycol) - 1))) {
  prob_matrix <- prob_byrow
} else {
  prob_matrix <- prob_bycol
}

stopifnot(mean(abs(rowSums(prob_matrix) - 1)) < 1e-6)

val_train$predicted_label <- factor(
  official_labels[max.col(prob_matrix, ties.method = "first")],
  levels = official_labels
)
val_train$prediction_confidence <- apply(prob_matrix, 1, max)
val_train$image_status <- "Training Image"

# --- C. Combine Samples ---
if (!"cell_labels" %in% names(colData(val_unseen))) {
  val_unseen$cell_labels <- "unlabelled"
}

shared_cols <- intersect(names(colData(val_train)), names(colData(val_unseen)))
colData(val_train) <- colData(val_train)[, shared_cols, drop = FALSE]
colData(val_unseen) <- colData(val_unseen)[, shared_cols, drop = FALSE]

val_spe <- cbind(val_train, val_unseen)
val_images <- c(sampled_train, sampled_unseen)

message("Training Images: ", paste(sampled_train, collapse = ", "))
message("Unseen Images: ", paste(sampled_unseen, collapse = ", "))
```

### 3. Load Spatial TIFFs and Masks

```{r load-spatial}
images_pattern <- paste0("^", val_images, "[.]tiff$", collapse = "|")

val_images_obj <- loadImages(file.path(base_path, "img"), pattern = images_pattern)
channelNames(val_images_obj) <- rownames(val_spe)

val_masks_obj <- loadImages(file.path(base_path, "masks"), pattern = images_pattern, as.is = TRUE)

names(val_images_obj) <- sub("\\.tiff$", "", names(val_images_obj))
names(val_masks_obj) <- sub("\\.tiff$", "", names(val_masks_obj))

img_metadata <- S4Vectors::DataFrame(
  sample_id = names(val_images_obj),
  status = ifelse(names(val_images_obj) %in% sampled_train, "Training Image", "Unseen Image")
)
S4Vectors::mcols(val_images_obj) <- S4Vectors::mcols(val_masks_obj) <- img_metadata
```

### 4. Validation 1A: The Full TME Map (Model Predictions)

```{r plot-full-outlines-predictions}
present_labels_pred <- unique(as.character(val_spe$predicted_label))
cur_colors_pred <- list(predicted_label = cell_type_colors[names(cell_type_colors) %in% present_labels_pred])

pdf(file.path(plot_dir, "01_Validation_1A_Predicted_TME.pdf"), width = 20, height = 10)
plotPixels(
  image = val_images_obj,
  mask = val_masks_obj,
  object = val_spe,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  outline_by = "predicted_label",
  colour = cur_colors_pred,
  thick = FALSE,
  legend = list(
    outline_by.labels.cex = 0.6,
    outline_by.legend.cex = 0.6,
    outline_by.title.cex = 0.8
  ),
  image_title = list(text = paste(mcols(val_masks_obj)$sample_id, "-", mcols(val_masks_obj)$status, "(PREDICTED)"), cex = 1)
)
dev.off()
```

### 5. Validation 1B: The Full TME Map (GROUND TRUTH)

```{r plot-full-outlines-gt}
train_images_obj <- val_images_obj[names(val_images_obj) %in% sampled_train]
train_masks_obj <- val_masks_obj[names(val_masks_obj) %in% sampled_train]
train_spe <- val_spe[, val_spe$sample_id %in% sampled_train]

present_labels_gt <- unique(as.character(train_spe$cell_labels))
cur_colors_gt <- list(cell_labels = cell_type_colors[names(cell_type_colors) %in% present_labels_gt])

pdf(file.path(plot_dir, "02_Validation_1B_GroundTruth_TME.pdf"), width = 20, height = 5)
plotPixels(
  image = train_images_obj,
  mask = train_masks_obj,
  object = train_spe,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  outline_by = "cell_labels",
  colour = cur_colors_gt,
  thick = FALSE,
  legend = list(
    outline_by.labels.cex = 0.6,
    outline_by.legend.cex = 0.6,
    outline_by.title.cex = 0.8
  ),
  image_title = list(text = paste(mcols(train_masks_obj)$sample_id, "- Training Image (GROUND TRUTH)"), cex = 1)
)
dev.off()
```

### 6. Validation 1C: High-Contrast Tumor Consistency Map

```{r plot-tumor-consistency}
val_spe$tumor_vs_all <- ifelse(val_spe$predicted_label == "Tumor", "Tumor", "Other")
tumor_contrast_colors <- list(tumor_vs_all = c(Tumor = "sienna4", Other = "white"))

pdf(file.path(plot_dir, "03_Validation_1C_HighContrast_Tumor.pdf"), width = 20, height = 10)
plotPixels(
  image = val_images_obj,
  mask = val_masks_obj,
  object = val_spe,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  outline_by = "tumor_vs_all",
  colour = tumor_contrast_colors,
  thick = FALSE,
  image_title = list(text = paste(mcols(val_masks_obj)$status, "- Predicted Tumor Architecture (Brown vs White)"), cex = 1)
)
dev.off()
```

### 7. Validation 2A: Tumor Overlay Logic (PREDICTED)

```{r plot-tumor-overlay-predicted}
tumor_markers <- intersect(c("Ecad", "panCK"), rownames(val_spe))
val_spe$is_tumor_pred <- ifelse(val_spe$predicted_label == "Tumor", "Tumor", "Other")
tumor_outline_colors <- list(is_tumor_pred = c(Tumor = "#FFFFFF60", Other = NA))

pdf(file.path(plot_dir, "04_Validation_2A_Predicted_Tumor_Overlay.pdf"), width = 20, height = 10)
plotPixels(
  image = val_images_obj,
  mask = val_masks_obj,
  object = val_spe,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  colour_by = tumor_markers,
  outline_by = "is_tumor_pred",
  colour = tumor_outline_colors,
  bcg = list(panCK = c(0, 4, 1), Ecad = c(0, 4, 1)), # Dialed back to 4
  thick = FALSE,
  image_title = list(text = paste(mcols(val_masks_obj)$status, "- Predicted Tumor Outlines"), cex = 1)
)
dev.off()
```

### 8. Validation 2B: Tumor Overlay Logic (GROUND TRUTH)

```{r plot-tumor-overlay-gt}
train_spe$is_tumor_gt <- ifelse(train_spe$cell_labels == "Tumor", "Tumor", "Other")
tumor_outline_colors_gt <- list(is_tumor_gt = c(Tumor = "#FFFFFF60", Other = NA))

pdf(file.path(plot_dir, "05_Validation_2B_GroundTruth_Tumor_Overlay.pdf"), width = 20, height = 5)
plotPixels(
  image = train_images_obj,
  mask = train_masks_obj,
  object = train_spe,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  colour_by = tumor_markers,
  outline_by = "is_tumor_gt",
  colour = tumor_outline_colors_gt,
  bcg = list(panCK = c(0, 4, 1), Ecad = c(0, 4, 1)), # Dialed back to 4
  thick = FALSE,
  image_title = list(text = paste(mcols(train_masks_obj)$sample_id, "- Training Image (GROUND TRUTH Tumor)"), cex = 1)
)
dev.off()
```

### 9. Validation 3: The `mregDC` Ruleset Check

```{r plot-mregdc-overlay}
val_spe$is_mregDC <- ifelse(val_spe$predicted_label == "mregDC", "mregDC", "Other")
mregdc_outline_colors <- list(is_mregDC = c(mregDC = "#FFFFFF60", Other = NA))

pdf(file.path(plot_dir, "06_Validation_3_Predicted_mregDC.pdf"), width = 20, height = 10)
plotPixels(
  image = val_images_obj,
  mask = val_masks_obj,
  object = val_spe,
  cell_id = "ObjectNumber",
  img_id = "sample_id",
  colour_by = tumor_markers,
  outline_by = "is_mregDC",
  colour = mregdc_outline_colors,
  bcg = list(panCK = c(0, 4, 1), Ecad = c(0, 4, 1)), # Dialed back to 4
  thick = FALSE,
  image_title = list(text = paste(mcols(val_masks_obj)$status, "- Predicted mregDC Outlines"), cex = 1)
)
dev.off()
```